---
title: "NIJ MetaT analysis"
format: pdf
---

```{r libs}
#| include: false
#adapted from: https://combine-australia.github.io/RNAseq-R/06-rnaseq-day1.html#Differential_expression_with_limma-voom
set.seed(3)
library(readxl)
library(writexl)
library(here)
library(ggplot2)
library(ggpubr)
library(MASS)
library(edgeR)
library(limma)
library(DESeq2)
library(kableExtra)
# library(reshape2)
library(pals)
library(cowplot)
library(ggpubr)
library(tidyverse)
library(knitr)

colpalette2= c("khaki3","steelblue3","orange","chocolate","aquamarine4","plum3","black","mediumpurple4","violetred3","brown4","chartreuse3","lavenderblush4","lightpink","mediumpurple3","lightcyan1","palegreen4","red1","wheat4","papayawhip","tomato3")
```

This script contains anaysis and visualizations of gene expression in decomposition soils from the NIJ-ARF study. This study followed n = 6 replicate human donors over one year of decomposition in the spring (n = 3) and wither (n = 3). We chose to focus on the spring study, and assess gene expression at study days 0, 12, 58, 86, 168, and 376.

```{r data}
#| include: false
# load counts matrix and metadata
NIJ_all_counts = read_xlsx(here::here("data/NIJ_rawcounts_matrix.xlsx")) #counts
gene_ref = NIJ_all_counts %>% select(Gene, Gene_length)
```

```{r}
#| include: false
# load sample metadata
metadata = read_csv(here::here("data/SP_recombined_soildata.csv")) #metadata
metadata$Sample_name = as.factor(metadata$Sample_name)
metadata$Trt = as.factor(metadata$Trt)
metadata$Rep = as.factor(metadata$Rep)
metadata$Study_day = factor(metadata$Study_day, ordered = TRUE, levels = c(0, 12, 58, 86, 168, 376))
metadata$Study_con = factor(metadata$Study_con, ordered = TRUE, levels = c("con", "d.0", "d.12", "d.58", "d.86", "d.168", "d.376"))
metadata$Stage = as.factor(metadata$Stage)

metadata_wide <- metadata %>% pivot_wider(names_from = parameter, values_from = value)

# load eggnog annotation file
gene_an = read_excel(here::here("data/NIJ_all_annotations.emapper.annotations.xlsx"), sheet = "R")
COG_cat = read_excel(here::here("data/NIJ_all_annotations.emapper.annotations.xlsx"), sheet = "COG_prim")
```


```{r count_an}
#| include: false
#| eval: false
#make a raw counts dataframe with annotations

NIJ_all_counts_ann <- NIJ_all_counts %>%
  rename(query = Gene) %>%
  left_join(gene_an, by = "query")

write_xlsx(NIJ_all_counts_ann, here::here("data/NIJ_rawcounts_annotated.xlsx"))
```

```{r GOI}
#| include: false
NIJ_all_counts_ann <- read_excel(here::here("data/NIJ_rawcounts_annotated.xlsx"))
# Genes of interest: 

## look for annamox genes via EC
  #hyradize synthase subunit - Hzs (EC: 1.7.2.7)
  Hzs = NIJ_all_counts_ann %>% filter(EC == "1.7.2.7")
  
  #hydrazine dehydrogenase - Hdh (EC: 1.7.2.8)
  Hdh = NIJ_all_counts_ann %>% filter(EC == "1.7.2.8")
```

## Assess Data Quality

We will pass our counts table to `DGEList()` to create a `DGElist` object, which is used by edgeR. This will allow us to normalize counts and perform differential analyses.

```{r edgeR_setup}
#| include: false
#format the data

#dataframe of count data only:
countdata = NIJ_all_counts %>% 
  column_to_rownames(var = "Gene") %>%
  select(-Gene_length)
  
#check colnames in counts file is in the same order as sample ID in metadata file
md <- metadata %>% pivot_wider(names_from = parameter, values_from = value)
all(colnames(countdata)==md$Sample_name)

#convert counts into DGElist object to be used by edgeR
y = DGEList(countdata)

names(y) #what is stored in y

trt = paste(md$Trt) #make an object for the variable trt
trt = factor(trt) #make sure trt is a factor
y$samples$trt = trt #add trt to the DGElist object under the samples tab

study_day = paste(md$Study_day)
study_day = factor(study_day)
y$samples$study_day = study_day

study_con = paste(md$Study_con)
study_con = factor(study_con)
y$samples$study_con = study_con

y$samples
```

```{r}
class(y)
```

There are `r nrow(y$counts)` genes in our dataset.

### Filter lowly expressed genes:

Lowly expressed genes provide little evidence for differential expression and can hinder some statistical analyses. They can also reduce power to detect deferentially expressed genes when FDR methods are applied. You can filter by counts per million threshold in n number of samples; might try TPM. From edgeR: rule of thumb is that a gene is required to have a count of 5-10 in a library to be considered for DE.

edgeR has the function `filterByExpr()`, which will keep rows (genes) that meet the threshold in a minimum number of samples. The minimum number of samples is determined by the group factor in the `DGElist`. edgeR will assess the smallest sample size within the groups defined and make this the threshold. If you added metadata under a name other than 'group', you can specify the experimental varible you want edgR to consider. I will use day as groups.

```{r remove_low}
#| include: false
#note: could add annotation information, especially if using entrezid; will just move on and add that info later (in the master enggnog annotation file)

keep = filterByExpr(y, group=study_con)
y.con = y[keep, , keep.lib.sizes=FALSE]
```

We started with `r nrow(y$counts)` genes. Filtering lowly expressed genes removed `r nrow(y$counts)-nrow(y.con$counts)`, meaning `r nrow(y.con$counts)` remain.

### Library sizes

```{r}
#| echo: false
#| warning: false
# check data quality
barplot(y.con$samples$lib.size, names=colnames(y.con),las=2, ann=FALSE, cex.names=0.75)
mtext(side =1, text="Sample", line = 4)
mtext(side = 2, text="Library size (millions)", line = 3)
title("Library sizes (raw counts)")
```

### log2 normalize counts

Count data is not normally distributed (this was expected), need to scale (log2) to look at the distribution of raw counts.

```{r}
#| echo: false
y.con.log = cpm(y.con,log=TRUE)

boxplot(y.con.log, xlab="", ylab="Log2 counts per million", las=2)
abline(h = median(y.con.log),col="purple")
title("Boxplots of logCPMs (unnormalised)")

saveRDS(y.con.log, here("data/NIJ_filter_log2.rds"))
```

Most samples are within the median line, only samples that may be suspect are SP1_376 and SP1_86.

### Normalize for composition bias

edgeR applies a TMM normalization to reduce bias between libraries. In this normalization approach, a set of normalization factors are calculated to define an effective library size. This uses a weighted trimmed mean of the log expression ratios between samples, correcting for sequencing depth, RNA composition, and gene length. Normalization factors multiply to unity across all libraries. Normalization factor < 1 for a library, means the library size will be scaled down (counts are scaled up), while a normalization factor > 1 means the library will be scale up (counts scaled down).

```{r}
#| include: false
y_norm = calcNormFactors(y.con)
y_norm$samples

saveRDS(y_norm, here("data/NIJ_DGEList_TMM.rds"))

y_norm_counts = cpm(y_norm) #extract TMM normalized counts; note while we are using the edgeR cpm function, this command will check the DGElist object for norm.factors that are produed when TMM normalization is completed using the calcNormFactors(); While not in this code, counts produced from the non-normalized df (y) and TMM norm (y_norm) were compared and do differ even though the $counts list within both DGElists are the same
```

*Compare log2 vs TMM normalized counts:*

log2 normalized:
```{r}
#| echo: false
plotMD(y.con.log, column = 5)
abline(h=0, col="grey")
```

TMM normalized:
```{r}
#| echo: false
plotMD(y_norm_counts, column = 5)
abline(h=0, col="grey")
```

```{r}
#| include: false
#| eval: false
y_norm_counts = y_norm_counts %>% 
  as.data.frame() %>%
  rownames_to_column("query") %>%
  left_join(gene_an, by = "query") %>%
  left_join(COG_cat, by = "COG_category")

write_xlsx(y_norm_counts, here::here("data/NIJ_TMM_countstable.xlsx"))
```

```{r}
#| include: false
y_norm_counts = read_xlsx(here::here("data/NIJ_TMM_countstable.xlsx"))
```

TPM normalized:
```{r TPM}
#| include: false
#try with TPM data:
tpm3 <- function(counts,len) {
  x <- counts/len
  return(t(t(x)*1e6/colSums(x)))
}

#first calculate TPM:
#gene lengths need to be in kbp
gene_ref = gene_ref %>% mutate(Gene_length_kbp = Gene_length/1000)

countdata_TPM = tpm3(countdata, gene_ref$Gene_length_kbp)
countdata_TPM = as.data.frame(countdata_TPM)
```

```{r}
#| echo: false
plotMD(countdata_TPM, column = 5)
abline(h=0, col="grey")
```

```{r}
#| include: false
#| eval: false
# prove the tpm3 function step by step
# RPK = countdata/gene_ref$Gene_length_kbp #one: divide counts by gene length (in kbp) 
# RPK_sf = colSums(RPK)/1000000
# TPM = RPK/RPK_sf #this results in the same values as those calculated by the tpm3 function

countdata_TPM <- countdata_TPM %>%
  rownames_to_column(var = "query") %>%
  left_join(gene_an, by = "query")

write_xlsx(countdata_TPM, here("data/NIJ_TPM_countstable.xlsx"))
```

```{r}
#| include: false

countdata_TPM <- read_xlsx(here("data/NIJ_TPM_countstable.xlsx"))

TPM_long <- countdata_TPM %>%
  pivot_longer(cols = starts_with(c("CON", "SP"))) 
```

## How do samples compare to each other (MDS)

```{r}
#| include: false
#multidimensional scaling plot by study day:

levels(metadata$Study_day) #how many study days and in what order are they stored
#phylum_colors <- c("#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD","#AD6F3B", "#673770","#D14285", "#652926", "#C84248", "#8569D5", "#5E738F","#D1A33D", "#8A7C64", "#599861")
col.day = c("#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD","#AD6F3B")[metadata$Study_day] #assign colors to each study day
data.frame(metadata$Study_day, col.day)
```


```{r}
#| echo: false
#redo the MDS with color by study day
plotMDS(y.con, col=col.day)

legend("topleft",
       fill = c("#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD"),
       legend = levels(metadata$Study_day))
title("Study Day")
```

Note: the distance between each pair of samples in the MDS plot are calculated and the leading fold change (root mean square of the largest 500 log2-fold changes) between that pair of samples.

### MDS with ggplot:

#### Colored by decomposition stage: 

```{r}
#| include: false
#tidy MDS plot with ggplot2
mds = plotMDS(y.con) #save output of plotMDS to an R object. This will extract the plotting values as a matrix that we can use to replot in ggplot2

mds_df <- data.frame("Sample_name" = rownames(mds$distance.matrix.squared),
                "Dim1" = mds$x,
                "Dim2" = mds$y) %>%
  left_join(metadata_wide, by = "Sample_name")

library(RColorBrewer)
brewer.pal(7, "Dark2")
#"#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E", "#E6AB02", "#A6761D"

x_lab <- paste0("LogFC Dim1 (", round(mds$var.explained[1] * 100, 2), "%)")
y_lab <- paste0("LogFC Dim2 (", round(mds$var.explained[2] * 100, 2), "%)")

mds_stage <- 
    ggplot(mds_df, aes(x=Dim1, y=Dim2, color= Stage, fill = Stage)) + geom_point(aes(color = Stage, shape = Rep), size = 5) + scale_shape_manual(values=c(15,16,17,18)) +
    scale_fill_manual(values = c("#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E", "#E6AB02", "#A6761D")) +
    scale_color_manual(values = c("#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E", "#E6AB02", "#A6761D")) +
    labs(x = x_lab, 
         y = y_lab,
         title = "Distance between gene expression\n across decomposition stages (MDS)") + 
    theme_bw()
```

```{r}
#| echo: false
#| warning: false
mds_stage
```

#### Colored by study day: 

```{r}
#| include: false
#| echo: false
#| warning: false
mds_df$day_num = as.numeric(as.character(mds_df$Study_day))
x_lab <- paste0("LogFC Dim1 (", round(mds$var.explained[1] * 100, 2), "%)")
y_lab <- paste0("LogFC Dim2 (", round(mds$var.explained[2] * 100, 2), "%)")

ggplot(mds_df, aes(x=Dim1, y=Dim2, color = Study_day, shape = Rep)) + geom_point(size = 5)+ scale_shape_manual(values=c(15,16,17,18)) + scale_color_manual(values = c("#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E", "#E6AB02", "#A6761D")) + labs(x = x_lab, y = y_lab, title = "Distance between gene expression\n over time (MDS)") + theme_bw()
```

```{r}
#| include: false
x_lab <- paste0("LogFC Dim1 (", round(mds$var.explained[1] * 100, 2), "%)")
y_lab <- paste0("LogFC Dim2 (", round(mds$var.explained[2] * 100, 2), "%)")

# colfunc <- colorRampPalette(c("lightblue","darkblue"))
# colfunc(6)
cols = c("#ADD8E6" ,"#8AACD3", "#6781C1", "#4556AF" ,"#222B9D", "#00008B")

#pink to dark blue pal
cols2 = c("#FB7B8E","#FAA7B8","#CDB3D4","#AAB6FB","#6096FD","#031B88")
cols2 <- setNames(cols2, levels(mds_df$Study_day)) 
cols2_df <- as.data.frame(cols2) %>% rownames_to_column(var = "Study_day")

mds_df2 <- mds_df %>%
  mutate(phase = case_when(Trt == "Control" ~ "Control",
                           Study_day == 0 ~ "Initial",
                           Study_day == 12 ~ "Bloom",
                           Study_day == 58 ~ "Climax",
                           Study_day == 86 ~ "Decline",
                           Study_day == 168 ~ "Recovery",
                           Study_day == 376 ~ "Recovery",
                           TRUE ~ NA_character_),
         phase = factor(phase, levels = c("Control", "Initial", "Bloom", "Climax", "Decline", "Recovery"))) 

# ,
#          color = case_when(Study_day == 0 ~ cols2[1],
#                            Study_day == 12 ~ cols2[2],
#                            Study_day == 58 ~ cols2[3],
#                            Study_day == 86 ~ cols2[4],
#                            Study_day == 168 ~ cols2[5],
#                            Study_day == 376 ~ cols2[6],
#                            TRUE ~ NA_character_)

cols3 = c("grey50", "#FB7B8E","#FAA7B8","#CDB3D4","#6096FD","#031B88")
cols3 <- setNames(cols3, levels(mds_df2$phase))

color_df <- mds_df2 %>%
  select(Study_day, phase) %>%
  distinct() %>% 
  left_join(cols2_df, by = "Study_day")


mds_day <- mds_df2 %>%
  ggplot(aes(x = Dim1, y = Dim2, shape = phase)) +
  geom_hline(yintercept= 0, color = "grey95") + 
  geom_vline(xintercept = 0, color = "grey95") + 
  geom_point(aes(color = Study_day, fill = Study_day), size = 5) +
  scale_shape_manual("Decomposition Phase", values = c(4, 15, 17, 16, 25, 18)) + 
  scale_color_manual("Study day", values = cols2) + 
  scale_fill_manual("Study day", values = cols2) +
  labs(x = x_lab, y = y_lab) + 
  theme_bw() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) 
        

saveRDS(mds_day, here::here("figures/MDS.rds"))
```

```{r}
#| echo: false
#| warning: false
mds_day <- readRDS(here::here("figures/MDS.rds"))
mds_day + labs(title = "Distance between gene expression\n over time (MDS)")
```

There is nice separation with time, and some overlap between 186 and 376. All controls and timepoint 0 samples cluster together. Final timepoint samples (376) do not cluster with their respective initial timepoint samples, suggesting gene expression is still altered after 1-year of decomposition. All but the control at study day 168 cluster together. This sample was take in October while all others were taken between May and July, suggesting seasonal differences may occur, but were not fully captured by the timepoints chosen for analysis. 

```{r}
#| include: false

png(here::here("figures/MDS_studyday.png"), units = "mm", width = 150, height = 100, res = 300)
mds_day
dev.off()

tiff(here::here("figures/MDS_studyday.tiff"), units = "mm", width = 150, height = 100, res = 300)
mds_day
dev.off()
```

## CCA plot - how do gene expression profiles correlate to environmental variables?

Environmental variables of interest:
ADH_ambient, ADH_internal, ADH_soil, Gravimetric_moisture, pH, EC, DO, CO2_μmol_gdw,  NH4_mg_gdw, NO3_mg_gdw, N_percent, C_percent, CN_ratio

```{r}
#| include: false

CCA_meta = metadata %>% 
  pivot_wider(names_from = parameter, values_from = value) %>%
  select(Sample_name, ADH_ambient, ADH_internal, ADH_soil, temp_ambient, temp_soil,
         Gravimetric_moisture, pH, EC, DO, CO2_umol_gdw,  NH4_mg_gdw, NO3_mg_gdw, N_percent, C_percent,
         CN_ratio)

#Canonical Correlation analysis (CCA) is a dimentionality reduction technique similar to PCA or SVD, however CCA deals with two high dimensional datasets at the same time and aims to find projections that account for most of co-variance between two datasets
  
TMM_count_table <- y_norm_counts %>%
  select(-seed_ortholog, -evalue, -score, -eggNOG_OGs, -max_annot_lvl, -COG_category,
         -Description, -Preferred_name, -GOs, -EC, -KEGG_ko, -KEGG_Pathway, -KEGG_Module,
         -KEGG_Reaction, -KEGG_rclass, -BRITE, -KEGG_TC, -CAZy, -BiGG_Reaction, -PFAMs, -Ann) %>%
  column_to_rownames(var = "query") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "Sample_name")
  
x_counts = TMM_count_table %>% select(-Sample_name)
y_meta = CCA_meta %>% select(-Sample_name)
```

```{r}
#| include: false
#use adonis to look for significant variables
library(vegan)
x.adonis <- adonis(x_counts ~ ., data=y_meta)
x.adonis$aov.tab #significant variables:  ADH_internal, soil temmp,pH, EC; ADH_ambient, ambient temp, ADH_soil < 0.2
cca_aov <- as.data.frame(x.adonis$aov.tab)

write.csv(cca_aov, here::here("data/allvar_cca_permanova.csv"), row.names = TRUE)

cca_aov_df <- cca_aov %>%
  rownames_to_column(var = "variable")

# write_xlsx(cca_aov, here::here("data/allvar_cca_permanova_aov.xlsx"))

#library(CCA)
#cc_results = cancor(x_counts,y_meta) Note: this is canonical correlation analysis, which differs from canonical correspondence analysis!

#extract environmental variables with p values < 0.2
bestEnv = rownames(x.adonis$aov.tab)[x.adonis$aov.tab$`Pr(>F)`<=0.2]

#last two are NA, need to remove
bestEnv = bestEnv[!is.na(bestEnv)]

r_all_cca <- cca(x_counts ~ ., data=y_meta)
r_all_cca
saveRDS(r_all_cca, here::here("data/cca_all"))

cca1 <- round(((r_all_cca$CCA$eig[[1]] / sum(r_all_cca$CCA$eig)) * 100), 1)
cca2 <- round(((r_all_cca$CCA$eig[[2]] / sum(r_all_cca$CCA$eig)) * 100), 1)
x_lab <- paste0("CCA1 (", cca1, "%)")
y_lab <- paste0("CCA2 (", cca2, "%)")


vif <- as.data.frame(vif.cca(r_all_cca))

scrs_all = scores(r_all_cca, display =c("sp","wa","lc","bp","cn"))
attributes(scrs_all) #check attributes

#extract site data first: this dataset contains one row for each observation and coordinates for CCA1 and CCA2. 
scrs_all$sites
df_all_sites <- data.frame(scrs_all$sites)
df_all_sites$Sample_name <- TMM_count_table$Sample_name
df_all_sites <- df_all_sites %>% left_join(metadata, by = "Sample_name")

#plot sites
p2 <- ggplot(df_all_sites) + 
  aes(CCA1, CCA2) +
  geom_hline(yintercept= 0, color = "grey95") + 
  geom_vline(xintercept = 0, color = "grey95") +
  geom_point(size = 5, aes(color = Study_day, shape = Rep)) +
  scale_shape_manual(values=c(15,16,17,18)) + 
  scale_color_manual(values = cols2) + 
  labs(x = x_lab, y = y_lab, color = "Study Day") +
  theme_bw()

multiplier = vegan:::ordiArrowMul(scrs_all$biplot)

labels <- c("ADH[A]", "ADH[I]", "ADH[S]", "T[A]", "T[S]", "GWC", "pH", "EC", "DO", "CO[2]", "NH[4]", "NO[3]", "N*'%'", "C*'%'", "C:N")
df_all_arrows <- as.data.frame(scrs_all$biplot*multiplier) %>% 
  rownames_to_column(var = "variable") %>%
  add_column(labels = labels) %>%
  left_join(cca_aov_df, by = "variable") %>%
  mutate(sig = if_else(`Pr(>F)` <= 0.05, "sig", "not_sig")) %>%
  mutate(CCA1 = if_else(CCA1 >= 0, CCA1 + 0.1, CCA1 - 0.1),
         CCA2 = if_else(CCA2 >= 0, CCA2 + 0.1, CCA2 - 0.1)) %>%
  mutate(CCA2 = if_else(variable == "EC", CCA2 + 0.09, CCA2))

text_labs <- as.data.frame(df_all_arrows)[, c("CCA1", "CCA2", "sig", "labels")]
text_labs  

p2 +
  geom_segment(data = df_all_arrows, aes(x = 0, y= 0, xend = CCA1, yend = CCA2), arrow = arrow(length = unit(0.2, "cm")), color = "black", alpha = 0.5) + 
  geom_text(data = text_labs[text_labs$sig == "sig", ], aes(CCA1, CCA2, label = text_labs[text_labs$sig == "sig", "labels"]), color = "red", parse = TRUE) +
  geom_text(data = text_labs[text_labs$sig == "not_sig", ], aes(CCA1, CCA2, label = text_labs[text_labs$sig == "not_sig", "labels"]), color = "black", parse = TRUE) + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())   
```


```{r}
#| include: false
p <- p + geom_text(data=as.data.frame(df_arrows*1.1), aes(CCA1, CCA2, label = rownames(df_arrows)), color="black")
p
```

```{r}
#| include: false

#cca with all chosen environmental parameters
y_meta <- CCA_meta %>% select(-Sample_name, -ADH_ambient, -ADH_internal, -ADH_soil)

x_sub_adonis <- adonis(x_counts ~ ., data=y_meta)
x_sub_adonis$aov.tab #significant variables:  ADH_internal, soil temmp,pH, EC; ADH_ambient, ambient temp, ADH_soil < 0.2
cca_sub_aov <- as.data.frame(x_sub_adonis$aov.tab)

write.csv(cca_sub_aov, here::here("data/cca_sub_permanova.csv"), row.names = TRUE)

r_sub_cca <- cca(x_counts ~ ., data = y_meta)
r_sub_cca
saveRDS(r_sub_cca, here::here("data/cca_subset"))

plot(r_sub_cca)

summary(r_sub_cca)

library(ggvegan)

p1 <- autoplot(r_sub_cca, layers = c("sites", "biplot"))

p1 + theme_bw()

cca1 <- round(((r_sub_cca$CCA$eig[[1]] / sum(r_sub_cca$CCA$eig)) * 100), 1)
cca2 <- round(((r_sub_cca$CCA$eig[[2]] / sum(r_sub_cca$CCA$eig)) * 100), 1)
x_lab <- paste0("CCA1 (", cca1, "%)")
y_lab <- paste0("CCA2 (", cca2, "%)")

t <- fortify(r_sub_cca) 

t %>% 
  filter(score == "sites") %>%
  mutate(Sample_name = TMM_count_table$Sample_name) %>% left_join(metadata_wide, by = "Sample_name")%>%
  mutate(phase = case_when(Trt == "Control" ~ "Control",
                           Study_day == 0 ~ "Initial",
                           Study_day == 12 ~ "Bloom",
                           Study_day == 58 ~ "Climax",
                           Study_day == 86 ~ "Decline",
                           Study_day == 168 ~ "Recovery",
                           Study_day == 376 ~ "Recovery",
                           TRUE ~ NA_character_),
         phase = factor(phase, levels = c("Control", "Initial", "Bloom", "Climax", "Decline", "Recovery"))) %>%
  ggplot() + 
  aes(CCA1, CCA2) +
  geom_hline(yintercept= 0, color = "grey95") + 
  geom_vline(xintercept = 0, color = "grey95") +
  geom_point(size = 5, aes(color = phase, shape = Rep)) +
  scale_shape_manual(values=c(15,16,17,18)) + 
  scale_color_manual(values = cols3) + 
  labs(x = x_lab, y = y_lab, color = "Phase") +
  theme_bw()

get_cca_arrows <- function(fortify) {
  want <- fortify$score == "biplot"
  vars <- ggvegan:::getDimensionNames(fortify)
  mul <- ggvegan:::arrow_mul(fortify[want, vars, drop = FALSE],
                             fortify[!want, vars, drop = FALSE])

  arrows <- fortify %>%
    filter(score == "biplot") %>%
    mutate(across(CCA1:CCA6, ~ .x * mul))

  return(arrows)
}

arrows <- get_cca_arrows(t) %>%
  mutate(labels = case_when(label == "temp_ambient" ~ "T[A]",
                            label == "temp_soil" ~ "T[S]",
                            label == "Gravimetric_moisture" ~ "GWC",
                            label == "pH" ~ "pH",
                            label == "EC" ~ "EC",
                            label == "DO" ~ "DO",
                            label == "CO2_umol_gdw" ~ "CO[2]",
                            label == "NH4_mg_gdw" ~ "NH[4]",
                            label == "NO3_mg_gdw" ~ "NO[3]",
                            label == "N_percent" ~ "N*'%'",
                            label == "C_percent" ~ "C*'%'",
                            label == "CN_ratio" ~ "C:N",
                            TRUE ~ NA_character_)) %>%
  rename(variable = label) %>%
  left_join(cca_aov_df, by = "variable") %>% # show full permanova results
  mutate(sig = if_else(`Pr(>F)` <= 0.05, "sig", "not_sig")) 
  # %>%
  # mutate(CCA1 = if_else(CCA1 >= 0, CCA1 + 0.1, CCA1 - 0.1),
  #        CCA2 = if_else(CCA2 >= 0, CCA2 + 0.1, CCA2 - 0.1)) %>%
  # mutate(CCA2 = if_else(variable == "EC", CCA2 + 0.09, CCA2))


arrows 

text_labs <- as.data.frame(arrows)[, c("CCA1", "CCA2", "sig", "labels")] %>%
  mutate(CCA1 = case_when(labels == "pH" ~ CCA1 - 0.1,
                          labels == "EC" ~ CCA1 + 0.15,
                          labels == "T[S]" ~ CCA1 + 0.13,
                          labels == "GWC" ~ CCA1 + 0.13,
                          labels == "C:N" ~ CCA1 - 0.15,
                          labels == "T[A]" ~ CCA1 - 0.1,
                          labels == "CO[2]" ~ CCA1 + 0.13,
                          labels == "N*'%'" ~ CCA1 + 0.1,
                          labels == "NH[4]" ~ CCA1 + 0.1,
                          labels == "C*'%'" ~ CCA1 + 0.075,
                          labels == "NO[3]" ~ CCA1 + 0.2,
                          TRUE ~ CCA1),
         CCA2 = case_when(labels == "pH" ~ CCA2 + 0.1,
                          # labels == "EC" ~ CCA2 - 0.04,
                          labels == "DO" ~ CCA2 + 0.13,
                          labels == "T[S]" ~ CCA2 - 0.13,
                          labels == "GWC" ~ CCA2 + 0.13,
                          labels == "C:N" ~ CCA2 - 0.03,
                          labels == "T[A]" ~ CCA2 - 0.1,
                          labels == "CO[2]" ~ CCA2 - 0.05,
                          labels == "N*'%'" ~ CCA2 - 0.1,
                          labels == "NH[4]" ~ CCA2 - 0.15,
                          labels == "C*'%'" ~ CCA2 - 0.1,
                          labels == "NO[3]" ~ CCA2 - 0.1,
                          TRUE ~ CCA2)) 

cca_p <- t %>% 
  filter(score == "sites") %>%
  mutate(Sample_name = TMM_count_table$Sample_name) %>% left_join(metadata_wide, by = "Sample_name")%>%
  mutate(phase = case_when(Trt == "Control" ~ "Control",
                           Study_day == 0 ~ "Initial",
                           Study_day == 12 ~ "Bloom",
                           Study_day == 58 ~ "Climax",
                           Study_day == 86 ~ "Decline",
                           Study_day == 168 ~ "Recovery",
                           Study_day == 376 ~ "Recovery",
                           TRUE ~ NA_character_),
         phase = factor(phase, levels = c("Control", "Initial", "Bloom", "Climax", "Decline", "Recovery"))) %>%
  ggplot() + 
  aes(CCA1, CCA2) +
  geom_hline(yintercept= 0, color = "grey95") + 
  geom_vline(xintercept = 0, color = "grey95") +
  geom_point(aes(color = Study_day, fill = Study_day, shape = phase), size = 5) +
  scale_shape_manual("Decomposition Phase", values = c(4, 15, 17, 16, 25, 18)) + 
  scale_color_manual("Study day", values = cols2) + 
  scale_fill_manual("Study day", values = cols2) +
  labs(x = x_lab, y = y_lab) +
  theme_bw() +
  geom_segment(data = arrows, aes(x = 0, y= 0, xend = CCA1, yend = CCA2), arrow = arrow(length = unit(0.2, "cm")), color = "black", alpha = 0.5) + 
  geom_text(data = text_labs[text_labs$sig == "sig", ], aes(CCA1, CCA2, label = text_labs[text_labs$sig == "sig", "labels"]), color = "red", parse = TRUE) +
  geom_text(data = text_labs[text_labs$sig == "not_sig", ], aes(CCA1, CCA2, label = text_labs[text_labs$sig == "not_sig", "labels"]), color = "black", parse = TRUE) + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) 
saveRDS(cca_p, here::here("figures/cca.rds"))
##############
# old

# cca1 <- round(((r_sub_cca$CCA$eig[[1]] / sum(r_sub_cca$CCA$eig)) * 100), 1)
# cca2 <- round(((r_sub_cca$CCA$eig[[2]] / sum(r_sub_cca$CCA$eig)) * 100), 1)
# x_lab <- paste0("CCA1 (", cca1, "%)")
# y_lab <- paste0("CCA2 (", cca2, "%)")


# # vif <- as.data.frame(vif.cca(r_sub_cca))

# scrs_all = scores(r_sub_cca, display =c("sp","wa","lc","bp","cn"))
# attributes(scrs_all) #check attributes

# #extract site data first: this dataset contains one row for each observation and coordinates for CCA1 and CCA2. 
# scrs_all$sites
# df_all_sites <- data.frame(scrs_all$sites)
# df_all_sites$Sample_name <- TMM_count_table$Sample_name
# df_all_sites <- df_all_sites %>% left_join(metadata_wide, by = "Sample_name")

# # library(ggfortify)
# # autoplot(r_sub_cca)

# #plot sites
# p2 <- df_all_sites %>%
#   mutate(phase = case_when(Study_day == 0 ~ "Initial",
#                            Study_day == 12 ~ "Bloom",
#                            Study_day == 58 ~ "Climax",
#                            Study_day == 86 ~ "Decline",
#                            Study_day == 168 ~ "Recovery",
#                            Study_day == 376 ~ "Recovery",
#                            TRUE ~ NA_character_),
#          phase = factor(phase, levels = c("Initial", "Bloom", "Climax", "Decline", "Recovery"))) %>%
#   ggplot() + 
#   aes(CCA1, CCA2) +
#   geom_hline(yintercept= 0, color = "grey95") + 
#   geom_vline(xintercept = 0, color = "grey95") +
#   geom_point(size = 5, aes(color = phase, shape = Rep)) +
#   scale_shape_manual(values=c(15,16,17,18)) + 
#   scale_color_manual(values = cols3) + 
#   labs(x = x_lab, y = y_lab, color = "Phase") +
#   theme_bw()

# #add biplot
# r_sub_cca$biplot
# # mul <- ggvegan:::arrowMul(r_sub_cca$biplot,
# #                           rbind(r_sub_cca$sites,
# #                                 r_sub_cca$species))
# # mul <- ggvegan:::arrowMul(cca.res$biplot,
# #                           rbind(cca.sites, cca.species))
# # cca.benthos <- data.frame(cca.res$biplot * mul)

# multiplier = vegan:::ordiArrowMul(scrs_all$biplot)

# library(ggfortify)
# ggvegan::autoplot(r_sub_cca)

# fortify(r_sub_cca)

# r_sub_ <- summary(r_sub_cca)
# ggvegan:::arrowMul(r_sub_$biplot,
#                           rbind(data.frame(r_sub_$sites), 
#                                 data.frame(r_sub_$species)))

#                           r_sub_cca$sites

# ub_aov_df <- cca_aov %>%
#   rownames_to_column(var = "variable")


# labels <- c("T[A]", "T[S]", "GWC", "pH", "EC", "DO", "CO[2]", "NH[4]", "NO[3]", "N*'%'", "C*'%'", "C:N")
# df_all_arrows <- as.data.frame(scrs_all$biplot*multiplier) %>% 
#   rownames_to_column(var = "variable") %>%
#   add_column(labels = labels) %>%
#   left_join(cca_aov_df, by = "variable") %>% # show full permanova results
#   mutate(sig = if_else(`Pr(>F)` <= 0.05, "sig", "not_sig")) %>%
#   mutate(CCA1 = if_else(CCA1 >= 0, CCA1 + 0.1, CCA1 - 0.1),
#          CCA2 = if_else(CCA2 >= 0, CCA2 + 0.1, CCA2 - 0.1)) %>%
#   mutate(CCA2 = if_else(variable == "EC", CCA2 + 0.09, CCA2))

# # these label positions are for Fig1.png below
# text_labs <- as.data.frame(df_all_arrows)[, c("CCA1", "CCA2", "sig", "labels")] %>%
#   mutate(CCA1 = case_when(labels == "pH" ~ CCA1 - 0.1,
#                           labels == "EC" ~ CCA1 + 0.15,
#                           labels == "T[S]" ~ CCA1 + 0.13,
#                           labels == "GWC" ~ CCA1 + 0.13,
#                           labels == "C:N" ~ CCA1 - 0.15,
#                           labels == "T[A]" ~ CCA1 - 0.1,
#                           labels == "CO[2]" ~ CCA1 + 0.13,
#                           labels == "N*'%'" ~ CCA1 + 0.1,
#                           labels == "NH[4]" ~ CCA1 + 0.1,
#                           labels == "C*'%'" ~ CCA1 + 0.075,
#                           labels == "NO[3]" ~ CCA1 + 0.2,
#                           TRUE ~ CCA1),
#          CCA2 = case_when(labels == "pH" ~ CCA2 + 0.1,
#                           # labels == "EC" ~ CCA2 - 0.04,
#                           labels == "DO" ~ CCA2 + 0.13,
#                           labels == "T[S]" ~ CCA2 - 0.13,
#                           labels == "GWC" ~ CCA2 + 0.13,
#                           labels == "C:N" ~ CCA2 - 0.03,
#                           labels == "T[A]" ~ CCA2 - 0.1,
#                           labels == "CO[2]" ~ CCA2 - 0.05,
#                           labels == "N*'%'" ~ CCA2 - 0.1,
#                           labels == "NH[4]" ~ CCA2 - 0.15,
#                           labels == "C*'%'" ~ CCA2 - 0.1,
#                           labels == "NO[3]" ~ CCA2 - 0.1,
#                           TRUE ~ CCA2)) 
# text_labs


# cca_p <- p2 +
#   geom_segment(data = df_all_arrows, aes(x = 0, y= 0, xend = CCA1, yend = CCA2), arrow = arrow(length = unit(0.2, "cm")), color = "black", alpha = 0.5) + 
#   geom_text(data = text_labs[text_labs$sig == "sig", ], aes(CCA1, CCA2, label = text_labs[text_labs$sig == "sig", "labels"]), color = "red", parse = TRUE) +
#   geom_text(data = text_labs[text_labs$sig == "not_sig", ], aes(CCA1, CCA2, label = text_labs[text_labs$sig == "not_sig", "labels"]), color = "black", parse = TRUE) + 
#   theme(panel.grid.major = element_blank(),
#         panel.grid.minor = element_blank()) 

# # + 
# #   geom_text(data = as.data.frame(df_all_arrows[df_all_arrows$sig == "sig", 2:3]*1.1), aes(CCA1, CCA2, label = df_all_arrows[df_all_arrows$sig == "sig", "labels"]), color="red", parse = TRUE) +
# #   geom_text(data = as.data.frame(df_all_arrows[df_all_arrows$sig == "not_sig", 2:3]*1.1), aes(CCA1, CCA2, label = df_all_arrows[df_all_arrows$sig == "not_sig", "labels"]), color="black", parse = TRUE)
```

```{r}
#| echo: false
#| warning: false
cca_p + labs(title = "CCA for NIJ ARF Spring")
```

```{r}
#| include: false
# width of 85 mm for half page width figure
# width of 170 mm for full page width figure
# maximum height of 225 mm for figure and legend
# image resolution of approximately 300 dpi (dots per inch) at the final size

png(here::here("figures/Fig1.png"), units = "mm", width = 170, height = 225, res = 300)
ggarrange(mds_day, cca_p, labels = c("A", "B"), nrow = 2, common.legend = TRUE, legend = "right")
dev.off()

tiff(here::here("figures/Fig1.tiff"), units = "mm", width = 170, height = 200, res = 300)
ggarrange(mds_day, cca_p, labels = c("A", "B"), nrow = 2, common.legend = TRUE, legend = "right")
dev.off()
```

```{r}
#| include: false

# write.csv(cca_aov, here::here("data/allvar_cca_permanova.csv"), row.names = TRUE)
# save all env variables PERMANOVA as a table
labels

rows <- c("ADH\\textsubscript{A}", "\\textbf{ADH\\textsubscript{I}}", "ADH\\textsubscript{S}", "T\\textsubscript{A}", "\\textbf{T\\textsubscript{S}}", "GWC", "\\textbf{pH}", "\\textbf{EC}", "DO", "CO\\textsubscript{2}", "NH\\textsubscript{4}", "NO\\textsubscript{3}", "N \\%", "C \\%", "C:N", "Residuals", "Total")

cca_aov_tbl <- read_csv(here::here("data/allvar_cca_permanova.csv")) %>%
  column_to_rownames(var = "...1")
rownames(cca_aov_tbl) <- rows

kbl(cca_aov_tbl,
    format = "latex",
    row.names = TRUE,
    digits = 3,
    booktabs = TRUE,
    escape = FALSE,
    col.names = c("Df", "SumOfSqs", "MeanSqs", "F Value", "R\\textsuperscript{2}", "Pr(>F)")) %>%
save_kable(here::here("tables/cca_permanova.pdf"), keep_tex = FALSE)


# rownames(mae_marker_anova_p)[1] <- "biologicalMarker"
# kbl(mae_marker_anova_p,
#     format = "latex",
#     row.names = TRUE,
#     # caption = "Model mean absoulte error (MAE) varies due to biological marker as reported by analysis of Variance (ANOVA). Model parameter biologicalMarker refers whether 16S, ITS, or both organisms were used in model construction.",
#     digits = 3,
#     booktabs = TRUE) %>% 
# save_kable(here::here("Mason_SoilMicrobe-PMI_PLoS_2024/figures/S4Table.pdf"), keep_tex = FALSE)
```

## Top 500 most variable genes (heatmaps)

```{r}
#| include: false
#heatmap.2() will calculate a matrix of euclidean distances from the logCPM for the 500 most variable genes

#estimate variance for each row in the log counts matrix
var_genes <- apply(y.con.log, 1, var)

#extract the top 500 MOST variable genes
select_var <- names(sort(var_genes, decreasing = TRUE))[1:500]

#use the top names to subset the counts matrix
highly_var_lcpm <- y.con.log[select_var,]
dim(highly_var_lcpm)

highly_var_lcpm_df <- highly_var_lcpm %>%
  as.data.frame() %>%
  rownames_to_column(var = "query") %>%
  left_join(gene_an, by = "query") %>%
  left_join(COG_cat, by = "COG_category")

highly_var_lcpm_df$Ann = factor(highly_var_lcpm_df$Ann, levels = c("Function Unknown","Lipid transport and metabolism", "Inorganic ion transport and metabolism","Cell wall/membrane/envelope biogenesis", "unclassified","Energy production and conversion" ,"Replication, recombination and repair","Post-translational modification, protein turnover, and chaperones", "Transcription","Signal transduction mechanisms" ,"Translation, ribosomal structure and biogenesis", "Cell cycle control, cell division, chromosome partitioning" ,"Amino acid transport and metabolism", "Cell motility", "Secondary metabolites biosynthesis, transport, and catabolism","Carbohydrate transport and metabolism"  ,"Intracellular trafficking, secretion and vesicular transport" ,"Nucleotide transport and metabolism","Coenzyme transport and metabolism"))

write_xlsx(highly_var_lcpm_df, here::here("data/NIJ_confilter_log_top500.xlsx"))
```

```{r}
#| include: false
highly_var_lcpm_df <- read_xlsx(here::here("data/NIJ_confilter_log_top500.xlsx"))

highly_var_lcpm_df_arran <- highly_var_lcpm_df %>%
  arrange(Ann)

highly_var_lcpm_df_arran$Ann = factor(highly_var_lcpm_df_arran$Ann, levels = c("Amino acid transport and metabolism", "Carbohydrate transport and metabolism", "Cell cycle control, cell division, chromosome partitioning", "Cell motility", "Cell wall/membrane/envelope biogenesis", "Coenzyme transport and metabolism", "Energy production and conversion", "Function Unknown", "Inorganic ion transport and metabolism", "Intracellular trafficking, secretion and vesicular transport", "Lipid transport and metabolism", "Nucleotide transport and metabolism", "Post-translational modification, protein turnover, and chaperones", "Replication, recombination and repair", "Secondary metabolites biosynthesis, transport, and catabolism", "Signal transduction mechanisms", "Transcription", "Translation, ribosomal structure and biogenesis", "unclassified"))
```

*What COG categories are present in the top 500 most variable expressed genes?*

```{r}
#| include: false
cog_cat_counts <- highly_var_lcpm_df %>%
  select(query, COG_category) %>%
  count(COG_category) %>%
  mutate(per = ((n/500)*100)) %>%
  left_join(COG_cat, by = "COG_category")

colpalette2 = c("khaki3","steelblue3","orange","chocolate","aquamarine4","plum3","black","mediumpurple4","violetred3","brown4","chartreuse3","lavenderblush4","lightpink","mediumpurple3","lightcyan1","palegreen4","red1","wheat4","papayawhip","tomato3")
nb.cols = 20
expanded_cols = colorRampPalette(colpalette2)(nb.cols)

top_500_cog_percent <- ggplot(cog_cat_counts, aes(x="", y=per, fill=Ann)) + 
  geom_bar(width = 1, stat = "identity") + 
  scale_fill_manual(values = expanded_cols) +
  labs(x = "", y = "Percentage of Genes") +
  theme_bw() + 
  guides(color = guide_legend(override.aes = list(size = 0.5))) +
  theme(legend.title = element_text(size = 8), 
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.8, "lines"))
```

```{r}
#| echo: false
#| warning: false
top_500_cog_percent
```

*How do the top 500 most variable expressed genes compare between samples?*

```{r}
#| include: false
library(gplots)
library(RColorBrewer)

#prep colors for the heatmap
mypalette <- brewer.pal(11, "RdYlBu")
morecols <- colorRampPalette(mypalette)
#colors for stage variable
brewer.pal(7, "Dark2")
col.study.con <- c("black","#FB7B8E","#FAA7B8", "#CDB3D4" ,"#AAB6FB", "#6096FD", "#031B88")[metadata_wide$Study_con]

col.COG = expanded_cols[highly_var_lcpm_df$Ann]
col.ann <- expanded_cols[highly_var_lcpm_df_arran$Ann]

samples <- str_subset(colnames(highly_var_lcpm_df_arran), "^(CON|SP.)")

highly_var_lcpm_plot <- highly_var_lcpm_df_arran %>%
  column_to_rownames(var = "query") %>%
  select(all_of(samples))

coords <- list("x" = 0.178, "y" = 1.91)
#plot heatmap - do not apply dendogram on y axis - order by COG cat
# heatmap.2(as.matrix(highly_var_lcpm_plot), 
#           col = rev(morecols(50)),
#           dendrogram = "none",
#           trace="none",
#           Rowv = FALSE,
#           RowSideColors = col.ann,
#           ColSideColors = col.study.con,
#           scale="row") +
#   legend(coords,      
#     legend = unique(metadata$Study_day),
#     col = unique(col.study.con), 
#     lty = 1,             
#     lwd = 5,           
#     cex =.7)

# coords <- list("x" = 0.178, "y" = 1.91)

highly_var_lcpm_df <- highly_var_lcpm %>%
  as.data.frame() %>%
  rownames_to_column(var = "query") %>%
  left_join(gene_an, by = "query") %>%
  left_join(COG_cat, by = "COG_category")


samples <- metadata_wide$Sample_name
highly_var_lcpm <- highly_var_lcpm_df %>%
  column_to_rownames(var = "query") %>%
  select(everything(samples)) %>%
  as.matrix()

#plot heatmap - dendogram on both sides
png(file = here::here("figures/Fig_S2.png"), width = 7, height = 8, units = "in", res = 300)
heatmap.2(highly_var_lcpm, 
          col=rev(morecols(50)),
          trace="none",
          ColSideColors = col.study.con,
          RowSideColors = col.COG,
          labRow = FALSE,
          scale="row")

legend(x = 0.2,
         y = 1.09,
         xpd=TRUE,
         legend = unique(metadata_wide$Study_day),
         col = unique(col.study.con), 
         title = "Study day",
         lty = 1,             
         lwd = 2,           
         cex =.7)
dev.off()
#          +
#   legend("topleft",
#          inset=c(1 , 1),
#          c("CON", levels(metadata_wide$Study_day)), 
#          col = unique(col.study.con),
#          ncol = 2,
#          pch=15,
#          title = "Day") 
# +
#   theme(legend.title = element_text(size = 8), 
#         legend.text = element_text(size = 8),
#         legend.key.size = unit(0.8, "lines"))
#   legend(coords,
#          legend = unique(metadata_wide$Study_day),
#          col = unique(col.study.con), 
#          lty = 1,             
#          lwd = 2,           
#          cex =.7)
```

## Gene counts over time

```{r}
#| include: false
#bar plot of gene counts of annotated (COG) genes over time
library(reshape2)
y_norm_counts_melt <- y_norm_counts %>%
  select(-seed_ortholog, -evalue, -score, -eggNOG_OGs, -max_annot_lvl, -Description, -Preferred_name, -GOs, -EC, -KEGG_ko, -KEGG_Pathway, -KEGG_Module, -KEGG_Reaction, -KEGG_rclass, -BRITE, -KEGG_TC, -CAZy, -BiGG_Reaction, -PFAMs) %>%
  melt(., id.vars = c("query", "COG_category", "Ann"))

y_norm_counts_sum <- y_norm_counts_melt %>%
  rename(Sample_name = variable) %>%
  summarise(sum_value = sum(value), .by = c("Sample_name", "Ann")) %>%
  left_join(metadata, by = "Sample_name")
```

### TMM counts by COG category

*Gene TMM for controls and each donor at each study day*

```{r}
#| echo: false
#| warning: false
colpalette2= c("khaki3","steelblue3","orange","chocolate","aquamarine4","plum3","black","mediumpurple4","violetred3","brown4","chartreuse3","lavenderblush4","lightpink","mediumpurple3","lightcyan1","palegreen4","red1","wheat4","papayawhip","tomato3")
nb.cols=25
expanded_cols=colorRampPalette(colpalette2)(nb.cols)

a = ggplot(y_norm_counts_sum, aes(x=Study_day, y= sum_value, fill = Ann)) + 
  geom_bar(stat = "identity") + scale_fill_manual(values = expanded_cols) + 
  theme(axis.text.x = element_text(hjust = 1, size=6)) + 
  scale_y_continuous(expand = c(0, 0)) + 
  guides(fill=guide_legend(ncol=1)) + 
  labs(x = "Study Day", y = "TMM Counts", fill = "COG Category") +
  facet_wrap(~Rep) +
  theme_bw()

addSmallLegend <- function(myPlot, pointSize = 3, textSize = 8, spaceLegend = 0.5) {
    myPlot +
        guides(shape = guide_legend(override.aes = list(size = pointSize)),
               color = guide_legend(override.aes = list(size = pointSize))) +
        theme(legend.title = element_text(size = textSize), 
              legend.text  = element_text(size = textSize),
              legend.key.size = unit(spaceLegend, "lines"))
}

addSmallLegend(a)
```

*Gene TMM for controls and each donor at each study day, but remove "unclassified"*

```{r}
#| echo: false
#| warning: false

#same plot, but remove any genes that were not assigned COG categories (aka 'unclassified')
y_norm_counts_sum_meta_COGhits = y_norm_counts_sum %>% filter(Ann != "unclassified")

b = ggplot(y_norm_counts_sum_meta_COGhits, aes(x=Study_day, y= sum_value, fill = Ann)) + geom_bar(stat = "identity") + scale_fill_manual(values = expanded_cols) + theme(axis.text.x = element_text(hjust = 1, size=6)) + scale_y_continuous(expand = c(0, 0)) + guides(fill=guide_legend(ncol=1))+ facet_wrap(~Rep) +theme_bw()

addSmallLegend(b)
```

*Gene TMM for controls and each donor at each study day, but remove "unclassified" and "Function Unkown"*

```{r}
#| echo: false
#| warning: false

#same plot, but remove any genes that were not assigned COG categories (aka 'unclassified') or function unknown (aka 'Function Unknown')
y_norm_counts_sum_meta_COGhits_notunknown = y_norm_counts_sum %>% filter(Ann != "unclassified",
                                                                         Ann != "Function Unknown")

c = ggplot(y_norm_counts_sum_meta_COGhits_notunknown, aes(x=Study_day, y= sum_value, fill = Ann)) + geom_bar(stat = "identity") + scale_fill_manual(values = expanded_cols) + theme(axis.text.x = element_text(hjust = 1, size=6)) + scale_y_continuous(expand = c(0, 0)) + guides(fill=guide_legend(ncol=1))+ facet_wrap(~Rep) +theme_bw()

addSmallLegend(c)
```

### Carbon, Oxygen, Hydroge, Nitrogen, and Sulfur gene expression over time

```{r}
#| include: false
#lets create a bubble plot of key carbon, oxygen, hydrogen, nitrogen, and sulfur genes (as found in Seyler et al 2021 Time series transcriptomics)

# set genes of interest
gene <- c("CBB cycle (cbbL)", "reverse TCA cycle (aclA)", "Wood-Ljungdahl pathway (cooS, acsA)", "methane/ammonia monooxygenase subunit A", "methane monooxygenase (pmoA)", "ubiquinol-cytochrome c reductase (UQRCRFS1, rip1, petA)", "cytochrome c oxidase (coxA/ccoN)", "cytochrome d ubiquinol oxidase (cydA)", "cytochrome o ubiquinol oxidase (cyoB)", "NiFe-hydrogenase (hyaB, hyaC)", "nitrogen fixation", "ammonia monooxygenase (amoA)", "nitrate reductase / nitrite oxidoreductase", "dissimilatory nitrate reductase (napA)", "dissimilatory nitrite reductase (cytochrome; nrfA)", "dissimilatory nitrite reductase (NADH; nirB)", "nitrite reductase (NO-forming; nirK)", "nitric oxide reductase (norB)", "nitrous-oxide reductase (nosZ)", "sulfide-quinone oxidoreductase (sqr)", "sulfur-oxidizing protein", "thiosulfate dehydrogenase (tsdA)", "dissimilatory sulfite reductase (dsrB)", "adenylylsulfate reductase (aprA)", "sulfate adenylyltransferase (sat, met3)", "assimilatory sulfite reductase (NADPH; cysl)", "assimilatory sulfite reductase (ferredoxin; sir)", "Hydroxylamine Dehydrogenase (HAO)")

EC_2018 <- c("4.1.1.39", "2.3.3.8", "1.2.7.4", "1.14.18.3,1.14.99.39", "1.14.18.3", "1.10.2.2", "1.9.3.1", "1.10.3.14", "1.10.3.10", "1.12.99.6", "1.18.6.1", "1.14.99.39", "1.7.5.1", "1.9.6.1", "1.7.2.2", "1.7.1.15", "1.7.2.1", "1.7.2.5", "1.7.2.4", "1.8.5.4", "NA", "1.8.2.2", "1.8.99.5", "1.8.99.2", "2.7.7.4", "1.8.1.2", "1.8.7.1", "1.7.2.6")

EC_2022 <- c("4.1.1.39", "2.3.3.8", "1.2.7.4", "1.14.18.3,1.14.99.39", "1.14.18.3", "7.1.1.8", "7.1.1.9", "7.1.1.7", "7.1.1.3", "1.12.99.6", "1.18.6.1", "1.14.99.39", "1.7.5.1", "1.9.6.1", "1.7.2.2", "1.7.1.15", "1.7.2.1", "1.7.2.5", "1.7.2.4", "1.8.5.4", "NA", "1.8.2.2", "1.8.99.5", "1.8.99.2", "2.7.7.4", "1.8.1.2", "1.8.7.1", "1.7.2.6")

EC_name <- c("cbbL", "aclA", "cooS, acsA", "pmoA - amoA", "pmoA", "rip1, petA", "coxA/ccoN", "cydA", "cyoB", "hyaB, hyaC", "nifH", "amoA", "nxrA, NarG", "napA", "nrfA", "nirB", "nirK", "norB", "nosZ", "sqr", "soxZ", "tsdA", "dsrB", "aprA", "sat", "cysI", "sir", "HOA")

element <- c("Carbon", "Carbon", "Carbon", "Carbon/Nitrogen", "Carbon", "Oxygen", "Oxygen", "Oxygen", "Oxygen", "H2", "Nitrogen", "Nitrogen", "Nitrogen", "Nitrogen", "Nitrogen", "Nitrogen", "Nitrogen", "Nitrogen", "Nitrogen", "Sulfur", "Sulfur", "Sulfur", "Sulfur", "Sulfur", "Sulfur", "Sulfur", "Sulfur", "Nitrogen")

GOI <- as.data.frame(cbind(gene, EC_2018, EC_2022, EC_name, element))
```

Genes of interest:

```{r}
GOI %>% kable()
```

```{r}
#| include: false
# extract counts for all GOI
GOI_long <- GOI %>%
  pivot_longer(cols = c("EC_2018", "EC_2022")) %>%
  rename(EC = value) %>% distinct(gene, EC_name, element, EC) %>%
  filter(EC != "NA")

GOI_counts_long <- y_norm_counts %>%
  select(-COG_category, -Ann, -seed_ortholog, -evalue, -score, -eggNOG_OGs,
         -max_annot_lvl, -Description, -GOs, -Preferred_name, -KEGG_ko, -KEGG_Pathway,
         -KEGG_Module, -KEGG_Reaction, -KEGG_rclass, -BRITE, -KEGG_TC, -CAZy, -BiGG_Reaction,
         -PFAMs) %>%
  pivot_longer(starts_with(c("CON", "SP"))) %>%
  filter(EC %in% GOI_long$EC) %>% 
  left_join(GOI_long, by = "EC") %>% #merge with GOI metadata list to take averages across queries of the same gene
# reformat the data: TMM counts for each gene and study day
  rename(Sample_name = name,
         TMM = value) %>%
  left_join(metadata, by = "Sample_name") %>%
  mutate(EC_name = as.factor(EC_name))

GOI_counts_long_sum <- GOI_counts_long %>%
  select(EC_name, Sample_name, Study_con, TMM) %>%
  group_by(Study_con, EC_name) %>%
  summarise(mean_TMM = mean(TMM)) %>% 
  pivot_wider(names_from = EC_name, values_from = mean_TMM, values_fill = 0)  %>%
  pivot_longer(cols = aclA:tsdA) %>%
  rename(EC_name = name) %>%
  left_join(GOI, by = "EC_name") %>%
  mutate(element = as.factor(element)) %>%
  mutate(EC_name = fct_reorder(EC_name, as.integer(element)))
  # mutate(Taxa = fct_reorder(Taxa, as.integer(Order)))

GOI_counts_long_sum$Study_con = factor(GOI_counts_long_sum$Study_con, levels=c("con","d.0","d.12","d.58","d.86","d.168","d.376"))

b = ggplot(GOI_counts_long_sum, aes(x = Study_con, y = EC_name)) + 
  geom_point(aes(size = value, fill = element), alpha = 0.75, shape = 21) +
  #scale_size_continuous(limits = c(0.000001, 100), range = c(1,17), breaks = c(1,10,50,75)) + 
  labs(x = "", y = "", size = "Average TMM", fill = "Element")  + 
  theme(legend.key=element_blank(), 
  axis.text.x = element_text(colour = "black", size = 12, face = "bold"), 
  axis.text.y = element_text(colour = "black", face = "bold", size = 11), 
  legend.text = element_text(size = 10, face ="bold", colour ="black"), 
  legend.title = element_text(size = 12, face = "bold"), 
  panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.2), 
  legend.position = "right") +  
  #scale_fill_manual(values = colours, guide = FALSE) + 
  scale_y_discrete(limits = rev(levels(GOI_counts_long_sum$EC_name))) +
  scale_x_discrete(labels = c("Control", "0", "12", "58", "86", "168", "376"))

```


```{r}
#| echo: false
addSmallLegend(b)
```

```{r}
#| include: false

png(here::here("figures/TPM_element_bubbleplot.png"), units = "mm", width = 170, height = 100, res = 300)
addSmallLegend(b)
dev.off()

tiff(here::here("figures/TPM_element_bubbleplot.tiff"), units = "mm", width = 170, height = 100, res = 300)
addSmallLegend(b)
dev.off()
```

### Lipase Genes

*How does expression of triacylglycerol lipase genes change over decomposition (determined by filter EC 3.1.1.3)*

```{r}
#| include: false

lipase_TPM <- TPM_long  %>%
  filter(EC == "3.1.1.3") %>%
  rename("Sample_name" = "name",
         "TPM" = "value") %>% 
  left_join(metadata_wide, by = "Sample_name") %>% #merge counts dataset with metadata 
  mutate(query = as.factor(query),
         tax = if_else(str_detect(max_annot_lvl, "Fungi"), "Fungi", "Bacteria"))

lipase_names = unique(lipase_TPM$query) #save an obeject with query names of all detected lipase genes, this can be used to pull out differential expression data

write_xlsx(lipase_TPM, here("data/lipase_TPM.xlsx"))
```

*All query genes with EC 3.1.1.3*

```{r}
#| echo: false
#average lipase gene transcript abundance ~ study day
lipase_TPM %>% 
  summarise(avg_TPM = mean(TPM), .by = c("Study_con", "query")) %>%
  ggplot() +
  aes(x = Study_con, y = avg_TPM, fill = "indianred") + 
  geom_boxplot() + 
  labs(title = "Mean TPM for each lipase gene at each study day", x = "Study Day", y = "Mean TPM Gene Counts") +
  theme_bw() +
  guides(fill="none")
```

*Sum lipase TPM in each sample*

```{r}
#| echo: false

lipase_TPM %>% 
  summarise(avg_TPM = mean(TPM),
            sum_TPM = sum(TPM),
            .by = c("Sample_name")) %>%
  left_join(metadata, by = "Sample_name") %>%
  ggplot() +
  aes(x = Study_con, y = sum_TPM, fill = "indianred") +
  guides(fill = "none") +
  geom_boxplot(position=position_dodge(1)) + 
  labs(title = "Sum triacylglycerol lipase TPM within each sample", x = "Study Day", y = "Sum TPM gene counts") +
  scale_x_discrete(labels = c("Control", "0", "12", "58", "86", "168", "376")) +
  theme_bw()
```

*Mean lipase TPM in each sample*

```{r}
#| echo: false

#lipase gene transcript abudance ~ study day and rep
lipase_TPM %>% 
  summarise(avg_TPM = mean(TPM),
            sum_TPM = sum(TPM),
            .by = c("Sample_name")) %>%
  left_join(metadata, by = "Sample_name") %>%
  ggplot() +
  aes(x = Study_con, y = avg_TPM, fill = "indianred") +
  guides(fill = "none") +
  geom_boxplot(position=position_dodge(1)) + 
  labs(title = "Mean TPM within each sample", x = "Study Day", y = "Mean TPM gene counts") +
  scale_x_discrete(labels = c("Control", "0", "12", "58", "86", "168", "376")) +
  theme_bw()
```


```{r}
#| include: false
#average lipase gene transcript abundance ~ donor ~ day with anova
lipase_TPM %>% 
  filter(Rep != "CON") %>% 
  summarise(avg_TPM = mean(TPM),
            .by = c("Study_day", "Rep")) %>%
  ggplot() +
  aes(x = Study_day, y = avg_TPM, fill = "indianred") +
  geom_boxplot() + 
  labs(x = "Study Day", y = "Mean TPM gene counts") +
  scale_x_discrete(labels = c("Control", "0", "12", "58", "86", "168", "376")) +
  stat_compare_means(method = "anova") + 
  guides(fill = "none") +
  theme_bw()
```

```{r}
#| echo: false
#average lipase gene transcript abundance ~ donor ~ day
l_tax <- lipase_TPM %>% 
  filter(Rep != "CON") %>% 
  summarise(avg_TPM = mean(TPM),
            .by = c("Study_day", "Rep", "tax")) %>%
  ggplot() + 
  aes(x = Study_day, y=avg_TPM, fill = "indianred") +
  geom_boxplot() +
  labs(x = "Study Day", y = "Mean TPM gene counts") +
  scale_x_discrete(labels = c("Control", "0", "12", "58", "86", "168", "376")) +
  facet_wrap(~tax) + 
  stat_compare_means(method="anova") +
  guides(fill = "none") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))

l_tax
```

Bacterial lipase gene expression decreases as decomposition progresses, while fungal lipase gene expression increases. 

```{r}
#| include: false
lipase_tax <- unique(lipase_TPM$max_annot_lvl)
```

Annotations for lipase genes are one of the following: `r lipase_tax`.

#### Fungal lipase genes

```{r}
#| echo: false
library(plotrix)
lipase_TPM %>% 
  filter(tax == "Fungi") %>%
  summarise(avg_TPM = mean(TPM),
            TPM_se = std.error(TPM),
            .by = c("Study_con", "query")) %>%
  ggplot() + 
  aes(x = Study_con, y = avg_TPM, group = query, color = query) + 
  geom_line(size = 1) + 
  geom_errorbar(aes(ymin = avg_TPM-TPM_se, ymax = avg_TPM+TPM_se), width = .2) + 
  labs(title = "Change in TPM over time for each identified fungal lipase gene",
       x = "Study Day",
       y = "Mean TPM Gene Counts",
       color = "Gene ID") +
  theme_bw()
```

```{r}
#| include: false
# other looks at the fungal lipase genes
lipase_TPM %>%
  filter(tax == "Fungi") %>%
  # filter(Rep %in% c("SP1","SP2","SP3")) %>%
  ggplot() +
  aes(x = Study_day, y = TPM, fill = "indianred") + 
  geom_boxplot() + 
  labs(title = "Fungal lipase gene expression below each donor", x = "Study Day", y = "TPM Gene Counts") +
  facet_wrap(~Rep) + 
  guides(fill = "none") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))
  #boxplot by study day faceted by donor/rep to assess variation

lipase_TPM %>% 
  filter(tax == "Fungi") %>%
  summarise(avg_TPM = mean(TPM),
            TPM_se = std.error(TPM),
            .by = c("Study_con", "query")) %>%
  ggplot() + 
  aes(x = Study_con, y = avg_TPM, fill = "indianred") + 
  geom_boxplot() + 
  guides(fill = "none") +
  theme_bw() + 
  stat_compare_means(method="anova") #plot average TPm of each fungal lipase gene at each study point

lipase_TPM %>% 
  filter(tax == "Fungi") %>%
  summarise(avg_TPM = mean(TPM),
            TPM_se = std.error(TPM),
            .by = c("Study_con", "query")) %>%
  ggplot() + 
  aes(x = Study_con, y = avg_TPM, fill = "indianred") + 
  geom_boxplot() + 
  guides(fill = "none") +
  theme_bw() + 
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = .2) #plot average TPm of each fungal lipase gene at each study point with dotplot
```

#### Bacterial lipase genes

```{r}
#| include: false
#bacterial lipase genes
lipase_TPM %>%
  filter(tax == "Bacteria") %>%
  summarise(avg_TPM = mean(TPM),
            TPM_se = std.error(TPM),
            .by = c("Study_con", "query")) %>%
  ggplot() + 
  aes(x = Study_con, y = avg_TPM, fill = "indianred") + 
  geom_boxplot() + 
  guides(fill = "none") +
  theme_bw() + 
  stat_compare_means(method = "anova") #plot average TPm of each bacterial lipase gene at each study point

lipase_TPM %>%
  filter(tax == "Bacteria") %>%
  ggplot() +
  aes(x = Study_day, y = TPM, fill = "indiandred") + 
  geom_boxplot() + 
  guides(fill = "none") +
  facet_wrap(~Rep) + 
  theme_bw() + 
  stat_compare_means(method="anova") #

a <- lipase_TPM %>%
  filter(tax == "Bacteria") %>%
  summarise(avg_TPM = mean(TPM),
            TPM_se = std.error(TPM),
            .by = c("Study_con", "query")) %>%
  ggplot() +
  aes(x = Study_con, y = avg_TPM, group = query, color = query) + 
  geom_line(size = 1) + 
  geom_errorbar(aes(ymin = avg_TPM-TPM_se, ymax = avg_TPM+TPM_se), width = .2) + 
  labs(title = "Change in TPM over time for each identified bacterial lipase gene", x = "Study Day", y = "Mean TPM Gene Counts") +
  scale_x_discrete(labels = c("Control", "0", "12", "58", "86", "168", "376")) +
  theme_bw() +
  theme(legend.position = "none")  
```

```{r}
#| echo: false
a
```

#### Stats

```{r}
#| include: false
# lipase stats
library(lme4)
library(lmerTest)

lipase_TPM_avg_day_rep <- lipase_TPM %>% 
  filter(Rep != "CON") %>% 
  summarise(avg_TPM = mean(TPM),
            .by = c("Study_day", "Rep", "tax")) 

lipase <-  lmer(avg_TPM ~ Study_day*tax+(1|Rep), data = lipase_TPM_avg_day_rep)
summary(lipase)

lipase_aov <- anova(lipase) 
write.csv(lipase_aov, here("data/lipase_TPM_full_anova.csv"))
# all, including interaction term significant: study day:tax; will split models by taxonomy (bacteria vs fungi)


################################
#####       Bacteria        ####
################################
lipase_bact <- lmer(avg_TPM ~ Study_day+(1|Rep), data = filter(lipase_TPM_avg_day_rep, tax == "Bacteria"))
summary(lipase_bact)

lipase_bact_anova <- anova(lipase_bact) # sig
write.csv(lipase_bact_anova, here("data/lipase_TPM_bact_anova.csv"))
lipase_bact_aov <- lipase_bact_anova

attach(filter(lipase_TPM_avg_day_rep, tax == "Bacteria"))
p_ttest = pairwise.t.test(avg_TPM, Study_day, p.adj = "bonf")
detach()
p_ttest_p = p_ttest$p.value

library(multcompView)
p_ttest_p_labels = data.frame(multcompLetters(p_ttest_p)['Letters']) 


################################
#####         Fungi         ####
################################
lipase_fung <- lmer(avg_TPM ~ Study_day+(1|Rep), data = filter(lipase_TPM_avg_day_rep, tax == "Fungi"))
summary(lipase_fung)

lipase_fung_aov <- anova(lipase_fung) #estimate: 0.005134; F = 4.509; p = 0.015
write.csv(lipase_fung_aov, here("data/lipase_TPM_fungi_anova.csv"))

attach(filter(lipase_TPM_avg_day_rep, tax == "Fungi"))
f_p_ttest = pairwise.t.test(avg_TPM, Study_day, p.adj = "bonf")
detach()
f_p_ttest_p = f_p_ttest$p.value
f_p_ttest_p_labels = data.frame(multcompLetters(f_p_ttest_p)['Letters']) 
```

Lipase full mod anova: 

```{r}
#| echo: false
kable(lipase_aov, digits = 3)
```

```{r}
#| include: false
#| eval: false

lipase_aov %>% 
  kbl(.,
      format = "latex",
      digits = 3,
      booktabs = TRUE) %>%
  save_kable(., file = here::here("figures/lipase_TPM_full_anova.pdf"))
```

Lipase - bacteria only mod anova:

```{r}
#| echo: false
#lipase - bacteria only mod anova
kable(lipase_bact_anova, digits = 3)
```

```{r}
#| include: false
#| eval: false

lipase_bact_anova %>% 
  kbl(.,
      format = "latex",
      digits = 3,
      booktabs = TRUE) %>%
  save_kable(., file = here::here("figures/lipase_TPM_bact_anova.pdf"))
```

Lipase - fungi only mod anova:

```{r}
#| echo: false
#lipase - fungi only mod anova
kable(lipase_fung_aov, digits = 3)
```

```{r}
#| include: false
#| eval: false

lipase_fung_aov %>% 
  kbl(.,
      format = "latex",
      digits = 3,
      booktabs = TRUE) %>%
  save_kable(., file = here::here("figures/lipase_TPM_fungi_anova.pdf"))
```

```{r}
#| include: false
lipase_TPM_sum_donor <- lipase_TPM %>%
  # filter(tax == "Fungi") %>%
  summarise(sum_TPM = sum(TPM),
            .by = c("Sample_name", "tax")) %>%
  left_join(metadata_wide, by = "Sample_name") %>%
  filter(Rep != "CON")

lipase_TPM_avg_donor <- lipase_TPM_sum_donor %>%
  summarise(avg_TPM = mean(sum_TPM),
            TPM_se = std.error(sum_TPM),
            .by = c("Study_day", "tax"))
  
lipase_TPM_sum_control <- lipase_TPM %>%
  summarise(sum_TPM = sum(TPM),
            .by = c("Sample_name", "tax")) %>%
  left_join(metadata_wide, by = "Sample_name") %>%
  filter(Rep == "CON") %>%
  rename(avg_TPM = sum_TPM)
```

*Fungal lipase patterns*

```{r}
#| echo: false
fl <- aov(lm(sum_TPM ~ Study_day, data = filter(lipase_TPM_sum_donor, tax == "Fungi"))) 
fl_t = TukeyHSD(fl)

#eztract p-values and labels
Tukey.levels <- fl_t$Study_day[,4]
Tukey.labels <- data.frame(multcompLetters(Tukey.levels)['Letters'])

#I need to put the labels in the same order as in the boxplot :
Tukey.labels$Study_day=rownames(Tukey.labels)
Tukey.labels$Study_day = factor(Tukey.labels$Study_day, ordered = TRUE, levels = c(0, 12, 58, 86, 168, 376))
Tukey.labels=Tukey.labels[order(Tukey.labels$Study_day) , ]

max(lipase_TPM_avg_donor$avg_TPM)

library(ggstar)

lipase_f <- ggplot(filter(lipase_TPM_avg_donor, tax == "Fungi")) + 
  aes(x = Study_day, y = avg_TPM, group = 1) + 
  geom_line(size = 1) + 
  geom_errorbar(aes(ymin = avg_TPM-TPM_se, ymax = avg_TPM+TPM_se), width = .2)  + 
  theme(legend.position = "none") + 
  geom_star(data = filter(lipase_TPM_sum_control, tax == "Fungi"), shape = 8, fill = "darkgoldenrod2", color = "darkgoldenrod2", size = 3) + 
  geom_text(data = Tukey.labels, aes(x = Study_day, y = 1.8, label = Letters) , size = 3) + 
  labs(x = "Study Day", y = "Mean TPM Gene Expression") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
lipase_f
```

*Bacterial lipase patterns*

```{r}
#| echo: false
bl <- aov(lm(sum_TPM ~ Study_day, data = filter(lipase_TPM_sum_donor, tax == "Bacteria"))) 
bl_t = TukeyHSD(bl)

#eztract p-values and labels
Tukey.levels <- bl_t$Study_day[,4]
Tukey.labels <- data.frame(multcompLetters(Tukey.levels)['Letters'])

#I need to put the labels in the same order as in the boxplot :
Tukey.labels$Study_day=rownames(Tukey.labels)
Tukey.labels$Study_day = factor(Tukey.labels$Study_day, ordered = TRUE, levels = c(0, 12, 58, 86, 168, 376))
Tukey.labels=Tukey.labels[order(Tukey.labels$Study_day) , ]

Tukey <- Tukey.labels %>%
  left_join(filter(lipase_TPM_avg_donor, tax == "Bacteria"), by = "Study_day") %>%
  mutate(ymax = avg_TPM + TPM_se)

aov <- paste0("Anova, p = ", round(lipase_bact_aov$`Pr(>F)`, 3))

lipase_b <- ggplot(filter(lipase_TPM_avg_donor, tax == "Bacteria")) + 
#  aes(x = Study_day, y = avg_TPM, group = 1) + 
  geom_line(aes(x = Study_day, y = avg_TPM, group = 1), size = 1) + 
  geom_errorbar(aes(x = Study_day, y = avg_TPM, ymin = avg_TPM-TPM_se, ymax = avg_TPM+TPM_se), width = .2)  + 
  theme(legend.position = "none") + 
  geom_star(data = filter(lipase_TPM_sum_control, tax == "Bacteria"), aes(x = Study_day, y = avg_TPM), fill = "darkgoldenrod2", color = "darkgoldenrod2", size = 3) + 
  geom_text(data = Tukey, aes(x = Study_day, y = ymax + 5, label = Letters) , size = 3) + 
  labs(x = "Study Day", y = "Mean TPM Gene Expression") +
  # annotate(geom = "text", x = 1.5, y = 95, label = aov, size = 2) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

lipase_b
```

#### How do fungal lipase trends correlate to budding yeast classes (*c__Saccharomycetes*, *c__Sordariomycetes*, *c__Eurotiomycetes*)?

```{r}
#| include: false
#| eval: false
#now lets extract Saccharomycetes, Sordariomycetes, and Eurotiomycetes in NIJ ARF soils
library(vegan)
library(scales)
library(grid)
library(phyloseq)

# First, create the variables for the imported data (first 3 are ITS, second 6 are 16S)
sharedfile <- here("data/NIJ_ARF_amplicon_data/NIJARFSP.trim.contigs.pcr.good.unique.precluster.pick.agc.shared")
taxfile <- here("data/NIJ_ARF_amplicon_data/NIJARFSP.trim.contigs.pcr.good.unique.precluster.pick.agc.0.05.cons.taxonomy")

metadata_seq_df <- read.csv(file = here("data/NIJ_ARF_amplicon_data/SP_ITS_metadata.csv"))

# Now, import the mothur data
mothur_data <- import_mothur(mothur_shared_file = sharedfile, mothur_constaxonomy_file = taxfile)

# import the metadata_seq file as a phyloseq object
metadata_seq <- sample_data(metadata_seq_df)
# In the metadata_seq file set Sample_name as the row name
rownames(metadata_seq) <- metadata_seq$Sample_name

# Merge metadata_seq file into phyloseq object created above
mothur_merged <- merge_phyloseq(mothur_data, metadata_seq)

mothur_sub = subset_samples(mothur_merged, Study_day %in% c(0, 12, 58, 86, 168, 376)) #only keep those samples associated with metatranscriptome study days

# Column names of taxonomy file are not particularly helpful, so we rename them.
colnames(tax_table(mothur_sub))
# Current names are "Rank 1",...through "Rank 7"
# Rename them to something more accessible:
colnames(tax_table(mothur_sub))<- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species") #ITS UNITE database

#Now let's prune the singletons and doubletons
mo_merge_pruned <- filter_taxa(mothur_sub, function (x) {sum(x > 0) > 2}, prune=TRUE)
mo_merge_pruned #Now we have 3289 taxa for SP

Class <- mo_merge_pruned %>%
  tax_glom(taxrank="Class") %>%
  transform_sample_counts(function(x) {x/sum(x)})
 
Class # 68 Classes

taxa <- c("c__Saccharomycetes", "c__Sordariomycetes", "c__Eurotiomycetes")

TOI <- subset_taxa(Class, Class %in% taxa)
TOI_table <- as(otu_table(TOI), "matrix")
if(taxa_are_rows(TOI)){TOI_table = t(TOI_table)}

TOI_table <- TOI_table %>%
  as.data.frame() %>%
  rename(c__Sordariomycetes = Otu00001, 
         c__Saccharomycetes = Otu00002,
         c__Eurotiomycetes = Otu00007) %>%
  rownames_to_column(var = "Sample_name") %>%
  left_join(metadata_seq_df, by = "Sample_name")


TOI_table$Study_day = factor(TOI_table$Study_day, ordered = TRUE, levels = c(0, 12, 58, 86, 168, 376))
```

```{r}
#| include: false
#| eval: false
write_xlsx(TOI_table, here("data/Fungal_lipase_taxa_abundance.xlsx"))
```

```{r}
#| include: false

TOI_table <- read_xlsx(here("data/Fungal_lipase_taxa_abundance.xlsx"))
metadata_wide <- metadata %>% pivot_wider(names_from = parameter, values_from = value)

# need to calculate relative values based on how we did DNA extractions (1/16th soil from interface, 15/16 from core)
TOI_adj_donor <- TOI_table %>%
  mutate(Rep = if_else(!is.na(str_split_i(Sample, pattern = "_", 3)), str_split_i(Sample, pattern = "_", 3), "CON")) %>%
  mutate(Rep = str_to_upper(Rep)) %>%
  filter(Rep != "CON") %>%
  select("c__Sordariomycetes", "c__Saccharomycetes", "c__Eurotiomycetes", "Depth", "Study_day", "Rep") %>% 
  pivot_wider(names_from = c("Study_day", "Rep"), values_from = c("c__Sordariomycetes", "c__Saccharomycetes", "c__Eurotiomycetes")) %>%
  column_to_rownames(var = "Depth") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "taxa_sample") %>%
  mutate(abundance = ((1/16)*interface) + ((15/16)*core)) %>%
  mutate(taxa_sample = str_split_i(taxa_sample, "__", 2)) %>%
  separate(taxa_sample, sep = "_", into = c("taxa", "Study_day", "Rep")) %>%
  select(-core, -interface) %>%
  pivot_wider(names_from = "taxa", values_from = "abundance") %>%
  rename(c__Sordariomycetes = Sordariomycetes,
         c__Saccharomycetes = Saccharomycetes,
         c__Eurotiomycetes = Eurotiomycetes) %>%
  inner_join(metadata_wide, by = c("Study_day", "Rep"))

# need to adjust controls (1/16th soil from interface, 15/16 from core)
TOI_adj_control <- TOI_table %>%
  mutate(Rep = if_else(!is.na(str_split_i(Sample, pattern = "_", 3)), str_split_i(Sample, pattern = "_", 3), "CON")) %>%
  mutate(Rep = str_to_upper(Rep)) %>%
  filter(Rep == "CON") %>%
  select("c__Sordariomycetes", "c__Saccharomycetes", "c__Eurotiomycetes", "Depth", "Study_day", "Rep") %>% 
  pivot_wider(names_from = c("Study_day", "Rep"), values_from = c("c__Sordariomycetes", "c__Saccharomycetes", "c__Eurotiomycetes")) %>%
  column_to_rownames(var = "Depth") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "taxa_sample") %>%
  mutate(abundance = ((1/16)*interface) + ((15/16)*core)) %>%
  mutate(taxa_sample = str_split_i(taxa_sample, "__", 2)) %>%
  separate(taxa_sample, sep = "_", into = c("taxa", "Study_day", "Rep")) %>%
  select(-core, -interface) %>%
  pivot_wider(names_from = "taxa", values_from = "abundance") %>%
  rename(c__Sordariomycetes = Sordariomycetes,
         c__Saccharomycetes = Saccharomycetes,
         c__Eurotiomycetes = Eurotiomycetes) %>%
  inner_join(metadata_wide, by = c("Study_day", "Rep"))


TOI_adj <- TOI_adj_donor %>%
  bind_rows(TOI_adj_control) %>%
  select(Sample_name, Study_day, Study_con, Rep, c__Sordariomycetes, c__Saccharomycetes, c__Eurotiomycetes) %>%
  pivot_longer(c( c__Sordariomycetes, c__Saccharomycetes, c__Eurotiomycetes)) %>%
  rename(abundance = value)
```

```{r}
#| include: false
ggplot(filter(TOI_adj, Rep != "CON")) + 
  aes(x = Study_day, y = abundance, fill = name) + 
  geom_col(position = position_dodge()) + 
  geom_star(data = filter(TOI_adj, Rep == "CON"), shape = 8, size = 3) + 
  scale_fill_manual(values = c("#882255", "#661100", "#6699CC")) + 
  theme_bw()

ggplot(filter(TOI_adj, Rep != "CON")) + 
  aes(x = Study_day, y = abundance, fill = name) + 
  geom_col(position = position_dodge()) + 
  scale_fill_manual(values = c("#882255", "#661100", "#6699CC")) + 
  theme_bw()
```

```{r}
#| echo: false
#| warning: false
# v %>%
#   mutate(names = str_remove(names, "c__"))
fl <- aov(lm(sum_TPM ~ Study_day, data = filter(lipase_TPM_sum_donor, tax == "Fungi"))) 
fl_t = TukeyHSD(fl)

#eztract p-values and labels
Tukey.levels <- fl_t$Study_day[,4]
Tukey.labels <- data.frame(multcompLetters(Tukey.levels)['Letters'])

#I need to put the labels in the same order as in the boxplot :
Tukey.labels$Study_day=rownames(Tukey.labels)
Tukey.labels$Study_day = factor(Tukey.labels$Study_day, ordered = TRUE, levels = c(0, 12, 58, 86, 168, 376))
Tukey.labels=Tukey.labels[order(Tukey.labels$Study_day) , ]

max(lipase_TPM_avg_donor$avg_TPM)
# get the maximum y value for tukey label placement
max_y <- TOI_adj_donor %>%
  bind_rows(TOI_adj_control) %>%
  select(Sample_name, Study_day, Study_con, Rep, c__Sordariomycetes, c__Saccharomycetes, c__Eurotiomycetes) %>%
  filter(Rep != "CON") %>%
  left_join(filter(lipase_TPM_avg_donor, tax == "Fungi"), by = "Study_day") %>%
  mutate(se_ymax = (avg_TPM + TPM_se)/2) %>%
  select(-tax, -avg_TPM, -TPM_se) %>%
  pivot_longer(c( c__Sordariomycetes, c__Saccharomycetes, c__Eurotiomycetes, se_ymax)) %>%
  group_by(Study_day) %>%
  summarise(max_y = max(value))

Tukey <- Tukey.labels %>%
  left_join(max_y, by = "Study_day")

#line plot of fungal lipase TPM over the three fungal classes relative abundance 
flp = filter(TOI_adj, Rep != "CON") %>%
  mutate(Study_day = factor(Study_day, levels = c("0", "12", "58", "86", "168", "376")),
         name = str_remove(name, "c__")) %>%
  ggplot() + 
  geom_col(aes(x = Study_day, y = abundance, fill = name), position=position_dodge()) +
  scale_fill_manual(values = c("#882255", "#661100", "#6699CC"))+
  geom_line(data = filter(lipase_TPM_avg_donor, tax == "Fungi"), aes(x = Study_day, y = avg_TPM/2, group = 1), size = 1) +
  geom_errorbar(data = filter(lipase_TPM_avg_donor, tax == "Fungi"), aes(x = Study_day, y = avg_TPM/2, group = 1, ymin = avg_TPM/2-TPM_se/2, ymax = avg_TPM/2+TPM_se/2), width = .2) +
  scale_y_continuous(limits = c(0, 1),
                     sec.axis = sec_axis(~.*2, name = "Mean TPM Gene Expression")) +
  geom_star(data = filter(lipase_TPM_sum_control, tax == "Fungi") , aes(x=Study_day, y=avg_TPM/2), shape = 8, fill = "darkgoldenrod2", color = "darkgoldenrod2", size = 3) +
  geom_text(data = Tukey, aes(x = Study_day, y = (max_y + 0.05), label = Letters), size = 3) + 
  # stat_compare_means(data = filter(lipase_TPM_avg_day_rep, tax == "Fungi"), aes(x = Study_day, y = avg_TPM), method = "anova", size = 2) +
  labs(x = "Study Day", y = "Relative Abundance", fill = "Class" ) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "inside",
        legend.justification = c("right", "top"),
        legend.background = element_rect(fill = NA))

lipase_f <- addSmallLegend(flp, textSize = 5) +
  theme(legend.title = element_blank())
lipase_f
```

```{r}
#| include: false
#| eval: false
lipase_TPM_fungi_TOI <- lipase_TPM %>%
  summarise(sum_TPM = sum(TPM),
            .by = c("Sample_name", "tax")) %>%
  left_join(metadata, by = "Sample_name") %>%
  filter(tax == "Fungi") %>%
  select(Sample_name, tax, sum_TPM) %>%
  left_join(TOI_adj_donor, by = "Sample_name")

library("Hmisc")
library("corrplot")

#be sure to remove any columns that are completely blank and that all variables are in numerical format
lipase_cor <- lipase_TPM_fungi_TOI %>% 
  filter(Rep != "CON") %>% 
  select(c__Saccharomycetes, c__Sordariomycetes, c__Eurotiomycetes, sum_TPM) %>%
  column_to_rownames(var = "Sample_name")

#make sure the dataset is a matrix
lipase_cor <- as.matrix(lipase_cor)

#run the rcorr command on the data matrix
NIJ_lipase_corr <- rcorr(lipase_cor, type="pearson")
signif(NIJ_lipase_corr$r,2)

#adjust the pvalue for multiple comparisions
pvaluesBH <- (p.adjust(NIJ_lipase_corr$P, method = "BH"))

NIJ_lipase_corr$P[]=p.adjust(NIJ_lipase_corr$P, method = "BH")
class(NIJ_lipase_corr$P)
NIJ_lipase_corr$P[]=p.adjust(NIJ_lipase_corr$P, method = "bonferroni")
p_adj=as.data.frame(signif(NIJ_lipase_corr$P,2))


#plot the correleation, where NS p-values appeat as white squares
corrplot(NIJ_lipase_corr$r, method=c("color"), tl.cex=0.8, tl.col="black", addgrid.col=c("gray"), sig.level=0.05)
```

```{r}
#| echo: false
  
bl <- aov(lm(sum_TPM ~ Study_day, data = filter(lipase_TPM_sum_donor, tax == "Bacteria"))) 
bl_t = TukeyHSD(bl)

#eztract p-values and labels
Tukey.levels <- bl_t$Study_day[,4]
Tukey.labels <- data.frame(multcompLetters(Tukey.levels)['Letters'])

#I need to put the labels in the same order as in the boxplot :
Tukey.labels$Study_day=rownames(Tukey.labels)
Tukey.labels$Study_day = factor(Tukey.labels$Study_day, ordered = TRUE, levels = c(0, 12, 58, 86, 168, 376))
Tukey.labels=Tukey.labels[order(Tukey.labels$Study_day) , ]

Tukey <- Tukey.labels %>%
  left_join(filter(lipase_TPM_avg_donor, tax == "Bacteria"), by = "Study_day") %>%
  mutate(ymax = avg_TPM + TPM_se)

aov <- paste0("Anova, p = ", round(lipase_bact_aov$`Pr(>F)`, 3))

lipase_b <- ggplot(filter(lipase_TPM_avg_donor, tax == "Bacteria")) + 
#  aes(x = Study_day, y = avg_TPM, group = 1) + 
  geom_line(aes(x = Study_day, y = avg_TPM, group = 1), size = 1) + 
  geom_errorbar(aes(x = Study_day, y = avg_TPM, ymin = avg_TPM-TPM_se, ymax = avg_TPM+TPM_se), width = .2)  + 
  theme(legend.position = "none") + 
  geom_star(data = filter(lipase_TPM_sum_control, tax == "Bacteria"), aes(x = Study_day, y = avg_TPM), fill = "darkgoldenrod2", color = "darkgoldenrod2", size = 3) + 
  geom_text(data = Tukey, aes(x = Study_day, y = ymax + 5, label = Letters) , size = 3) + 
  labs(x = "Study Day", y = "Mean TPM Gene Expression") +
  # annotate(geom = "text", x = 1.5, y = 95, label = aov, size = 2) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

lipase_b

```

```{r}
#| echo: false
#| warning: false
# lipase_b and lipase_f
ggarrange(lipase_b, lipase_f,  
          ncol = 2, 
          labels = c("A", "B"), 
          align = "v")
```

```{r}
#| include: false
# width of 85 mm for half page width figure
# width of 170 mm for full page width figure
# maximum height of 225 mm for figure and legend
# image resolution of approximately 300 dpi (dots per inch) at the final size

png(here::here("figures/Fig4.png"), units = "mm", width = 170, height = 100, res = 300)
ggarrange(lipase_b, lipase_f,  
          ncol = 2, 
          labels = c("A", "B"), 
          align = "v")
dev.off()

tiff(here::here("figures/Fig4.tiff"), units = "mm", width = 170, height = 100, res = 300)
ggarrange(lipase_b, lipase_f,  
          ncol = 2, 
          labels = c("A", "B"), 
          align = "v")
dev.off()
```

### Nitrogen Cycling Genes


*Nitrogen cycling genes*

```{r HAO}
#| include: false
#EC 1.7.2.6 - HAO
library(plotrix)

#look at TPM of HAO genes (determined by filter EC 1.7.2.6)
HAO_TPM <- TPM_long %>%
  filter(EC == "1.7.2.6") %>%
  rename(Sample_name = name,
         TPM = value) %>%
  left_join(metadata, by = "Sample_name") %>%
  mutate(query = factor(query),
        day = as.numeric(as.character(Study_day)))

#HAO gene transcript abudance ~ study day and rep
HAO_TPM %>%
  ggplot() +
  aes(x = Study_day, y = TPM) +
  geom_boxplot() + 
  theme_bw()

#average HAO gene transcript abundance ~ donor ~ day
HAO_TPM_avg_day = HAO_TPM %>% 
  filter(Rep != "CON") %>% 
  summarise(sum_TPM = sum(TPM),
            avg_TPM = mean(TPM),
            TPM_se = std.error(TPM), 
            .by = c(Study_day, Rep))

ggplot(HAO_TPM_avg_day, aes(x = Study_day, y = avg_TPM, fill = "indianred")) +
  geom_boxplot(position = position_dodge(1)) + 
  theme_bw() + 
  theme(legend.position = "none") +
  stat_compare_means(method = "anova")

ggplot(HAO_TPM_avg_day, aes(x = Study_day, y = sum_TPM, fill = "indianred")) +
  geom_boxplot(position = position_dodge(1)) + 
  theme_bw() + 
  theme(legend.position = "none") +
  stat_compare_means(method="anova")
```

```{r}
#| include: false
hao_lmer_df <- HAO_TPM %>%
  summarise(sum = sum(TPM), .by = c("Study_day", "Rep")) %>%
  filter(Rep != "CON")

HAO_mod <- lmer(sum ~ Study_day + (1|Rep), data = hao_lmer_df)
summary(HAO_mod)

write.csv(as.data.frame(anova(HAO_mod)), here("data/hao_lmer_anova.csv"))
```

```{r}
#| include: false
#calculate the average TPM of each gene at each study day (+con) for plot
HAO_TPM_avg = HAO_TPM %>% 
  summarise(avg_TPM = mean(TPM),
            TPM_se = std.error(TPM), 
            .by = c(Study_con, query)) 
  

ggplot(HAO_TPM_avg, aes(x = Study_con, y = avg_TPM, fill = "indianred")) + 
  geom_boxplot() + 
  theme_bw() + 
  stat_compare_means(method = "anova") #plot average TPm of each bacterial HAO gene at each study point

# library(lme4)
# library(lmerTest)
 
# # stats
# HAO_bact = lmer(avg_TPM ~ Study_day+(1|Rep), data = HAO_TPM_avg_day)
# summary(HAO_bact)

# HAO_bact_aov = anova(HAO_bact) 

# HAO_bacts = lmer(sum_TPM ~ Study_day+(1|Rep), data = HAO_TPM_avg_day)
# summary(HAO_bacts)

# HAO_bacts_aov = anova(HAO_bacts) 
```

```{r}
#| echo: false
ggplot(HAO_TPM_avg_day, aes(x = Study_day, y = sum_TPM, fill = "indianred")) +
  geom_boxplot(position=position_dodge(1)) + 
  labs(y = "TPM", x = "Study Day", title = "HAO gene expression over time") +
  guides(fill = "none") +
  theme_bw() + 
  stat_compare_means(method = "anova")
```


```{r}
#| include: false
n_genes <- GOI %>%
  filter(element == "Nitrogen") 

#average gene transcript abundance ~ donor ~ day
ngenes_TPM <- TPM_long %>%
  filter(EC %in% n_genes$EC_2018) %>%
  rename(Sample_name = name,
         TPM = value) %>%
  left_join(md, by = "Sample_name")

library(plotrix)

ngenes_TPM_day_avg <- ngenes_TPM %>%
  summarise(sum_TPM = sum(TPM),
            avg_TPM = mean(TPM),
            se_TPM = std.error(TPM),
            .by = c("EC.x", "Study_day", "Rep")) %>%
  rename(EC_2018 = EC.x) %>%
  left_join(n_genes, by = "EC_2018") %>%
  filter(Rep != "CON") %>%
  mutate(EC_name = factor(EC_name))
```

```{r}
#| echo: false
ggplot(ngenes_TPM_day_avg, aes(x = Study_day, y = sum_TPM, fill="indianred")) + 
  geom_boxplot() +
  labs(y = "TPM", x = "Study Day", "Expression of N-cycling genes over time") +
  facet_wrap(~EC_name, scales = "free_y") + 
  theme_bw() + 
  guides(fill = "none") +
  stat_compare_means(method="anova", size = 2, vjust = 1) + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))
```

```{r}
#| include: false
#| warning: false
### Controls
ngenes_TPM_avg_con <- ngenes_TPM %>%
  group_by(EC.x, Study_day, Rep) %>%
  summarise(sum_TPM = sum(TPM)) %>%
  rename(EC_2018 = EC.x) %>%
  left_join(n_genes, by = "EC_2018") %>%
  filter(Rep == "CON") %>%
  mutate(EC_name = factor(EC_name))

nc = ggplot(ngenes_TPM_avg_con) +
  aes(x = Study_day, y = sum_TPM, group = EC_name, color = EC_name) +
  labs(x = "Study Day", y = "TPM", color = "Gene Name") +
  geom_line(size = 1) + 
  scale_color_manual(values = colpalette2) +  
  theme_bw()+ 
  # theme(legend.position = "None") + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(limits = c(0, 190), breaks = c(0, 50, 100, 150))

nc 

### Decomp soils
ngenes_TPM_avg_decomp <- ngenes_TPM_day_avg %>%
  ungroup() %>%
  group_by(EC_2018, Study_day) %>%
  summarise(avg_TPM = mean(sum_TPM),
            TPM_se = std.error(sum_TPM)) %>%
  left_join(n_genes, by = "EC_2018") %>%
  mutate(EC_name = factor(EC_name))

nd = ggplot(ngenes_TPM_avg_decomp) +
  aes(x = Study_day, y = avg_TPM, group = EC_name, color = EC_name) +
  labs(x = "Study Day", y = "Mean TPM", color = "Gene Name") +
  geom_line(size = 1) + 
  scale_color_manual(values = colpalette2) + 
  theme_bw() + 
  geom_errorbar(aes(ymin = avg_TPM-TPM_se, ymax = avg_TPM+TPM_se), width = .2) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

nd
```


```{r}
#| include: false
# magnitude change
#look at general magnitude change in HAO, norB, and nosZ

HOA <- ngenes_TPM %>%
  filter(EC.x %in% n_genes[n_genes$EC_name == "HOA", "EC_2018"]) %>%
  group_by(Study_day, Rep) %>%
  summarise(sum_TPM = sum(TPM)) 

HOA_c <- HOA %>%
  filter(Rep == "CON") %>%
  summarise(con_avg = mean(sum_TPM))

HOA_d <- HOA %>%
  filter(Rep != "CON") %>%
  ungroup() %>%
  group_by(Study_day) %>%
  summarise(avg = mean(sum_TPM)) %>%
  left_join(HOA_c, by = "Study_day") %>%
  mutate(diff = avg - con_avg,
         times = avg / con_avg)
```


```{r}
#| echo: false
#| warning: false
ggarrange(nc, nd, labels = c("A", "B"), common.legend = TRUE)
```

```{r}
#| include: false
png(here::here("figures/Fig5.png"), units = "mm", width = 130, height = 100, res = 300)
ggarrange(nc, nd, labels = c("A", "B"), common.legend = TRUE)
dev.off()

tiff(here::here("figures/Fig5.tiff"), units = "mm", width = 130, height = 100, res = 300)
ggarrange(nc, nd, labels = c("A", "B"), common.legend = TRUE)
dev.off()
```

### Aminopeptidases

```{r}
#| include: false
#aminopeptidases: 3.4.11.X

aminopep_TPM <- TPM_long %>%
  filter(str_detect(EC, "3.4.11")) %>%
  rename(Sample_name = name,
         TPM = value) %>%
  left_join(metadata, by = "Sample_name") %>%
  mutate(query = factor(query),
         day = as.numeric(as.character(Study_day)))
```

```{r}
#| echo: false
#| warning: false
#lapase gene transcript abudance ~ study day and rep
ggplot(aminopep_TPM) + 
  aes(x = Study_day, y = TPM, fill = Rep) +
  geom_boxplot(position=position_dodge(1)) + 
  labs(title = "Aminopeptidases (EC 3.4.11.X) expression over time") +
  theme_bw()
```

*what genes are the outliers in the boxplot?*

```{r}
#| include: false
# what genes are the outliers in the boxplot?
outliers <- aminopep_TPM %>%
#  filter(Study_day == "0") %>%
  filter(TPM >= 5) %>%
  pull(query) %>%
  unique()

aminopep_TPM_outliers <- aminopep_TPM %>%
  filter(query %in% outliers) %>%
  mutate(names = str_c(str_split_i(max_annot_lvl, pattern = "\\|", 2), " - ", Preferred_name))
  
```

```{r}
#| echo: false
#| warning: false
ggplot(aminopep_TPM_outliers) +
  aes(x = Study_day, y = TPM, group = Rep, color = Rep) +
  geom_line() +
  facet_wrap(~names, scales = "free_y") +
  theme_bw()
```


```{r}
#| include: false
# I want to know the variance across all 1913 aminopeptidase genes
aminopep_TPM_avg_day_rep <- aminopep_TPM %>%
  filter(Rep != "CON") %>%
  group_by(Study_day, Rep) %>%
  summarise(avg_TPM = mean(TPM),
            se_TPM = std.error(TPM))
```

```{r}
#| include: false
ggplot(aminopep_TPM_avg_day_rep) +
  aes(x = Study_day, y = avg_TPM, group = Rep, color = Rep) +
  geom_line() + 
  geom_errorbar(aes(ymin = avg_TPM-se_TPM, ymax = avg_TPM+se_TPM), width=.2) +
  theme_bw()
```


### Differential Expression: Treatment

```{r}
#| include: false
#Differential expression with limma-voom
#limma includes voom, which will transform read counts into logCPMs while also accounting for mean-variance in the data
y_norm <- readRDS(here("data/NIJ_DGEList_TMM.rds"))
y.con.log <- readRDS(here("data/NIJ_filter_log2.rds"))

# metadata_wide <- metadata %>% pivot_wider(names_from = parameter, values_from = value)
trt = paste(metadata_wide %>% pull(Trt)) #make an object for the variable trt
trt = factor(trt) #make sure trt is a factor

#specify a design matrix without an intercept term (variable options: trt, study day, study con)

#design 1: which genes are differentially expressed in decomposition soil compared to control soils
design_trt = model.matrix(~ 0 + trt)
colnames(design_trt) = levels(trt) #make column names of design matrix clean

#Voom transform the data (edgeR TMM normalized df):
v = voom(y_norm, design_trt, plot = TRUE) #results in voom normalized log2 counts

#weights tab is the mean variance trends 

#compare normalized data to normalized data:
par(mfrow=c(1,2))
boxplot(y.con.log, xlab = "", ylab = "Log2 counts per million",las = 2,main = "Unnormalised logCPM")
abline(h = median(y.con.log), col = "blue") ## Let's add a blue horizontal line that corresponds to the median logCPM
boxplot(v$E, xlab = "", ylab = "Log2 counts per million", las = 2, main = "Voom transformed logCPM")
abline(h = median(v$E), col = "blue") ## Let's add a blue horizontal line that corresponds to the median logCPM
par(mfrow=c(1,1))

#test for DE:
  #1: fit linear model for each gene
fit = lmFit(v) #lmFit estimates group means according to the specified design matrix (here trt)

  #2: specify comparisons for test (which groups do we want to compared differences for?) Here: control compared to decomposition
cont.matrix = makeContrasts(trt = Control - Decomposition, levels = design_trt) #note: we can specify as many comparisons as we want in one command

  #3: Apply comparisons to the fit model; this will result in statistics and estimated parameters for the comparision(s)
fit.cont = contrasts.fit(fit, cont.matrix)

  #4: Perform empirical Bayes shrinkage on variances and estimate t-stats/p-values
fit.cont = eBayes(fit.cont)

dim(fit.cont)

#summary of DE:
summa.fit = decideTests(fit.cont)
trt_sum = summary(summa.fit) #in theory up should be upreg in control and down should be downreg in control

#make a dataframe of the normalized log2 expression values for all query genes used in DE analysis
trt_nlog2 <-  as.data.frame(v$E) %>%
  rownames_to_column(var = "query") %>%
  pivot_longer(-query, names_to = "Sample_name", values_to = "nlog2") %>%
  left_join(metadata_wide, by = "Sample_name") 
```

Number of genes whose expression was higher in the control samples (Up), lower in control samples (Down), or not significantly different.

```{r}
#| echo: false
kable(trt_sum)
```

`r trt_sum[which(rownames(trt_sum) == "Up"), "trt"]` genes were upregulated in control soils, while `r trt_sum[which(rownames(trt_sum) == "Down"), "trt"]` were downregulated in control soils.

```{r}
#| include: false
names <- data.frame("query" = rownames(fit.cont$coefficients)) %>%
  left_join(select(gene_an, query, Preferred_name), by = "query") %>%
  mutate(Preferred_name = ifelse(Preferred_name == "-", query, Preferred_name))
```

*Control vs decomposition MD plot*

```{r}
#| echo: false
#Control vs decomposition
plotMD(fit.cont, coef = 1, status = summa.fit[,"trt"], values = c(-1,1), hl.col=c("blue","red"))
```


*Volcano plot: top #100 of genes highlighted*
```{r}
#| echo: false
volcanoplot(fit.cont,coef = 1, highlight = 100, main="trt")
```

*Volcano plot: top #10 of genes highlighted*
```{r}
#| echo: false
volcanoplot(fit.cont,coef = 1, highlight = 10, main = "Control vs. Decomposition", names = row.names(fit.cont$coefficients))
```

```{r}
#| include: false
trt_contrast = as.data.frame(summa.fit) %>%
  rownames_to_column(var = "query") %>%
  rename(trt_cont = trt) 
#this is a dataframe of all the genes used for DE analysis; cell values can only be one of the following: -1, 0, or 1: -1 = downregulated, 0 = not significantly DE, 1 = upregulated

#save coefficients contrasts:
trt_coeff <- as.data.frame(fit.cont$coefficients) %>%
  rownames_to_column(var = "query") %>%
  rename(trt_coeff = trt) 
#extract coefficients for each comparison; these are the log2 fold change values for each gene

#save DE p-values
trt_p <- as.data.frame(fit.cont$p.value) %>%
  rownames_to_column(var = "query") %>%
  rename(trt_p = trt) #note: these p-values are the same as those reported in the F.p.value 


trt_DE_summary = trt_contrast %>%
  left_join(trt_coeff, by = "query") %>% 
  left_join(trt_p, by = "query") %>% 
  mutate(log_p = -log10(trt_p)) %>%
  left_join(gene_an, by = "query") %>% 
  left_join(COG_cat, by = "COG_category")

saveRDS(trt_DE_summary, here("data/DE_trt_summary.rds"))
```

*Top 6 upregulated genes in control soils:*

```{r}
#| echo: false
#top 6 upregulated genes in control soils
trt_DE_summary <- readRDS(here("data/DE_trt_summary.rds"))

trt_DE_summary %>%
  arrange(desc(trt_coeff)) %>%
  head() %>%
  select(query, trt_cont, trt_coeff, trt_p, max_annot_lvl, Preferred_name, Ann) %>%
  kable()
```

```{r}
#| include: false

# ordered by largest coefficient and p-value
trt_15_sig_up = trt_DE_summary %>%
arrange(desc(log_p), desc(trt_coeff)) %>%
top_n(15, log_p) #top 10 significant and upregulated genes in control soils

trt_DE_summary %>%
  filter(query == "k123_2470867_1")

# k123_2470867_1
```

```{r}
#| include: false
top_up = trt_DE_summary %>%
  top_n(1, trt_coeff) %>% 
  pull(query)
```

The gene with the largest coefficient was `r top_up`. The expression in all samples, by treatment type is below.

```{r}
#| echo: false
library(RColorBrewer)
library(glue)
nice.col = brewer.pal(7, name="Dark2")
title <- glue("{top_up}",
              ": ",
              "{gene_an %>% filter(query == top_up) %>% pull(Description)}")

stripchart(v$E[top_up,] ~ trt,vertical=TRUE, 
           las=2, cex.axis = 0.8, pch = 16, cex = 1.3, 
           col = nice.col, method = "jitter", 
           ylab = "Normalized log2 expression", 
           main = title)
```

The gene with the largest coefficient, and smallest p-value was `r head(trt_15_sig_up, 1) %>% pull(query)`. The expression in all samples, by treatment is below. 

```{r}
#| echo: false
top_s_up <- head(trt_15_sig_up, 1) %>% pull(query)
title <- glue("{top_s_up}",
              ": ",
              "{gene_an %>% filter(query == top_s_up) %>% pull(Description)}}")


stripchart(v$E[top_s_up,] ~ trt, vertical=TRUE, 
           las=2, cex.axis = 0.8, pch = 16, cex = 1.3, 
           col = nice.col, method = "jitter", 
           ylab = "Normalized log2 expression", 
           main = title)
```

Multiple genes for secretion system proteins are in the top significantly upregulated genes, including type II and multiple type IV. Seems like signalling/cell motility might be downregulated in decomp - kind of makes sense as nutrients are increasing.

```{r}
#| include: false
#downregulated genes
head(trt_DE_summary %>% arrange(desc(log_p),trt_coeff)) #top 6 significant and downregulated genes in control soils

trt_15_sig_down <- trt_DE_summary %>% 
arrange(desc(log_p), trt_coeff) %>%
filter(trt_cont == "-1") %>% 
top_n(15, log_p) #top 10 significant and downregulated genes in control soils

trt_top_up.down = rbind(trt_15_sig_up, trt_15_sig_down)
```

A majority of the genes that are down regulated in control soils (therefore display higher expression in decomposition soils are related to heat shock response): 11/15 genes.


Differentially expressed genes between controls

```{r}
#| echo: false
#plot differentially expressed genes between controls
t <- trt_DE_summary %>% filter(trt_cont == "1") %>% filter(trt_p <= 0.05) %>% arrange(desc(trt_coeff)) %>% top_n(20, trt_coeff)
diff(trt_sig_up.20$query, t$query)
trt_sig_up.20 = trt_DE_summary %>% arrange(desc(log_p))%>% filter(trt_cont == "1") %>%top_n(20, log_p) #top 10 significant and downregulated genes in control soils



td <- trt_DE_summary %>% filter(trt_cont == "-1") %>% filter(trt_p <= 0.05) %>% slice_min(n = 20, order_by = trt_coeff)
trt_sig_down.20 = trt_DE_summary %>% arrange(desc(log_p))%>% filter(trt_cont == "-1") %>%top_n(20, log_p) #top 10 significant and downregulated genes in control soils

trt_top_up.down.20 = rbind(trt_sig_up.20, trt_sig_down.20)
write.csv(trt_top_up.down.20, here::here("data/top_40_DE_trt.csv"), row.names = FALSE)

length(unique(trt_sig_up.20$Ann)) #10 unique COGs
length(unique(trt_sig_down.20$Ann)) #4 unique COGs


#horizontal bar plot: 
trt_top_up.down.20$query = factor(trt_top_up.down.20$query)
trt_top_up.down.20$Ann = factor(trt_top_up.down.20$Ann, levels = c("Cell motility","Energy production and conversion","Transcription","Cell cycle control, cell division, chromosome partitioning","Carbohydrate transport and metabolism","Intracellular trafficking, secretion and vesicular transport","Signal transduction mechanisms","Translation, ribosomal structure and biogenesis","Post-translational modification, protein turnover, and chaperones", "Function Unknown", "unclassified"))
trt_top_up.down.20$max_annot_lvl = factor(trt_top_up.down.20$max_annot_lvl)

trt <- trt_top_up.down.20 %>%
  ggplot() +
  aes(x = query, y = trt_coeff, fill = Ann) + 
  geom_col(width = 0.5) +
  scale_fill_manual(values = expanded_cols) + 
  labs(y = "log2 Fold Change", 
      x = "Query Transcript",
      fill = "COG Category") +
  theme_bw() + 
  coord_flip() + 
  theme(axis.text.y = element_blank(),  #remove y axis labels
        axis.ticks.y = element_blank(),
        # axis.text.y = element_text(size = 6), 
        axis.text.x = element_text(size = 6),
        aspect.ratio = 2/1)

addSmallLegend_bottom <- function(myPlot, pointSize = 3, textSize = 8, spaceLegend = 0.5) {
    myPlot +
        guides(shape = guide_legend(override.aes = list(size = pointSize)),
               color = guide_legend(override.aes = list(size = pointSize)),
               fill = guide_legend(ncol = 1)) +
        theme(legend.title.position = "top",
              legend.title = element_text(size = textSize), 
              legend.text  = element_text(size = textSize),
              legend.key.size = unit(spaceLegend, "lines"),
              legend.position = "bottom")
}

addSmallLegend_bottom(trt)

trt_tax <- trt_top_up.down.20 %>%
  ggplot() +
  aes(x = query, y = trt_coeff, fill = max_annot_lvl) + 
  geom_col(width = 0.5) +
  scale_fill_manual(values = unname(kelly())) +
  labs(y = "log2 Fold Change", 
     x = "Query Transcript",
     fill = "Taxonomy") +
  theme_bw() + 
  coord_flip() + 
  theme(axis.text.y = element_blank(),  #remove y axis labels
        axis.ticks.y = element_blank(),
        # axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6), 
        aspect.ratio = 2/1)

addSmallLegend_bottom(trt_tax)


plot_grid(addSmallLegend_bottom(trt), addSmallLegend_bottom(trt_tax), ncol = 2, labels = "AUTO", align = "h")
```

```{r}
#| include: false

png(here::here("figures/Trt_DE_top20_barplot.png"), units = "mm", width = 170, height = 225, res = 300)
ggarrange(addSmallLegend_bottom(trt), addSmallLegend_bottom(trt_tax), ncol = 2, labels = "AUTO", align = "h")
dev.off()

tiff(here::here("figures/Trt_DE_top20_barplot.tiff"), units = "mm", width = 170, height = 225, res = 300)
ggarrange(addSmallLegend_bottom(trt), addSmallLegend_bottom(trt_tax), ncol = 2, labels = "AUTO", align = "h")
dev.off()
```

```{r}
#| include: false
#| eval: false
trt_top_up.down.20 %>% 
mutate(tax = str_split_i(max_annot_lvl, "\\|", 2)) %>%
select(query, trt_coeff, trt_p, tax, COG_category, Description, Preferred_name, EC) %>%
  kbl(.,
      format = "latex",
      digits = 3,
      col.names = c("Query", "Coefficient", "p-Value", "Taxonomic Annotation", "COG Category", "Description", "Gene Name", "EC"),
      booktabs = TRUE) %>%
  column_spec(6, width = "10em") %>%
  save_kable(., file = here::here("figures/Top20_DE_genes_trt.pdf"))
```

#### Heat Shock Proteins


```{r}
#| echo: false


# hs = c("k123_4681706_1","k123_2129273_1","k123_2784358_1","k123_2453574_1","k123_2580989_1","k123_784972_1","k123_1897324_1","k123_1119824_1","k123_3023451_1","k123_1018246_1","k123_4142344_1")
# all COG O in top 20 are heat shock/stress proteins 
hs <- trt_sig_down.20 %>%
  filter(COG_category == "O") %>%
  pull(query)

#average fold change of hsp
hs_fc = trt_DE_summary %>% 
  filter(query %in% hs) 

top_hsp <- hs_fc %>% 
  distinct(Preferred_name) %>% 
  pull(Preferred_name)
  
hs_fc %>% 
  summarise(avg_fc = mean(trt_coeff))

# white banners
library(plotrix)
trt_nlog2 %>% 
  left_join(gene_an, by = ("query")) %>%
  filter(query %in% hs) %>%
  summarize(mean = mean(nlog2),
            se = std.error(nlog2),
            .by = c("Preferred_name", "Study_day", "Rep")) %>%
  ggplot() +
  aes(x = Study_day, y = mean, group = Rep, color = Rep) + 
  geom_point(size = 1) +
  geom_line() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.1) +
  scale_color_manual(values = nice.col) + 
  facet_wrap( ~ Preferred_name) + 
  labs(title = "Expression of Top Downregulated Heat Shock Genes")+ 
  xlab("Study Day") + 
  ylab("Normalized log2 Expression") + 
  theme_bw() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.95, 0.05),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))

  # ggplot() +
  # aes(x = Study_day, y = nlog2, group = Rep, color = Rep) + 
  # geom_point(size =1 ) +
  # geom_line() +
  # scale_color_manual(values = nice.col) + 
  # facet_wrap( ~ query) + 
  # labs(title = "Expression of Top Downregulated Heat Shock Genes")+ 
  # xlab("Study Day") + 
  # ylab("Normalized log2 Expression") + 
  # theme_bw() +
  # theme(legend.position = "inside",
  #       legend.position.inside = c(0.9, 0.15),
  #       panel.grid.major = element_blank(),
  #       panel.grid.minor = element_blank(),
  #       strip.background = element_rect(fill = "white"))
```

```{r}
#| include: false

png(here::here("figures/Trt_DE_top15_mean_heatshockTPM.png"), units = "mm", width = 170, height = 225, res = 300)
trt_nlog2 %>% 
  filter(query %in% hs) %>%
  left_join(gene_an, by = ("query")) %>%
  summarize(mean = mean(nlog2),
            se = std.error(nlog2),
            .by = c("Preferred_name", "Study_day", "Rep")) %>%
  ggplot() +
  aes(x = Study_day, y = mean, group = Rep, color = Rep) + 
  geom_point(size = 1) +
  geom_line() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.1) +
  scale_color_manual(values = nice.col) + 
  facet_wrap( ~ Preferred_name) + 
  xlab("Study Day") + 
  ylab("Normalized log2 Expression") + 
  theme_bw() +
  # guides(color = guide_legend(ncol = 2)) +
  theme(legend.position = "inside",
        legend.position.inside = c(0.85, 0.2),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))
dev.off()

png(here::here("figures/Trt_DE_top15_heatshockTPM.png"), units = "mm", width = 170, height = 225, res = 300)
trt_nlog2 %>% 
  filter(query %in% hs) %>%
  ggplot() +
  aes(x = Study_day, y = nlog2, group = Rep, color = Rep) + 
  geom_point(size =1 ) +
  geom_line() +
  scale_color_manual(values = nice.col) + 
  facet_wrap( ~ query) + 
  xlab("Study Day") + 
  ylab("Normalized log2 Expression") + 
  theme_bw() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.9, 0.15),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white")) 
dev.off()

tiff(here::here("figures/Trt_DE_top15_heatshockTPM.tiff"), units = "mm", width = 170, height = 225, res = 300)
trt_nlog2 %>% 
  filter(query %in% hs) %>%
  ggplot() +
  aes(x = Study_day, y = nlog2, group = Rep, color = Rep) + 
  geom_point(size =1 ) +
  geom_line() +
  scale_color_manual(values = nice.col) + 
  facet_wrap( ~ query) + 
  xlab("Study Day") + 
  ylab("Normalized log2 Expression") + 
  theme_bw() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.9, 0.15),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white")) 
dev.off()
```

General TMM heat shock proteins for each donor and controls:

```{r}  
#| echo: false
# same heat shock protiens in TMM
phylum_colors <- c("#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD","#AD6F3B", "#673770","#D14285", "#652926", "#C84248", "#8569D5", "#5E738F","#D1A33D", "#8A7C64", "#599861")

y_norm_counts %>%
  pivot_longer(cols = starts_with(c("CON", "SP")), names_to = "Sample_name", values_to = "TMM") %>%
  filter(query %in% hs) %>%
  summarise(sum_TMM = sum(TMM), .by = c("Sample_name")) %>%
  left_join(metadata, by = "Sample_name") %>%
  ggplot() +
  aes(x = Study_day, y = sum_TMM, color = Rep) + 
  geom_point(size = 1) +
  scale_color_manual(values = phylum_colors) + 
  labs(y = "Gene Transcript Counts (TMM)",
       x = "Study Day",
       title = "Gene Transcript Counts (TMM) for heat shock\n protiens in the top 20 Downregulated Genes in Control Soils") +
  theme_bw()
```

```{r}
#| include: false
#determine magnitude of difference in exression for each query hs gene
y_norm_counts %>%
  pivot_longer(cols = starts_with(c("CON", "SP")), names_to = "Sample_name", values_to = "TMM") %>%
  filter(query %in% hs) %>%
  left_join(metadata, by = "Sample_name") %>%
  summarise(mean = mean(TMM), .by = c("Trt", "Study_day", "query")) %>%
  pivot_wider(names_from = Trt, values_from = mean) %>%
  mutate(diff = Decomposition - Control,
         times = Decomposition / Control)
  
```